---
title: "Engine + ModelProvider scaffold"
maturity: "validated"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Engine + ModelProvider scaffold}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette demonstrates the core architecture:

- `Entity` holds mutable state, sparse history, and an event log.
- `Engine` orchestrates the simulation **and materializes models** using a provider + model spec.
- `ModelBundle` supplies the dynamics.
- `ModelProviders` load bundles from different sources.


### Determinism, tie-breaking, and optional entity tags

When multiple processes propose events at the same `time_next`, the engine selects one deterministically by ordering on:

1. earliest `time_next`, then
2. lexicographic `process_id`.

This means `process_id` is not just a label, it is part of the model's deterministic resolution policy. If your model has a preferred priority when ties occur, encode that priority into the `process_id` (for example, a zero-padded numeric prefix).

For downstream linking in a HIPAA-conscious way, a `Entity` may also carry an optional single-column identifier `id` (default `NULL`). This is intended for *de-identified* tags (for example, a study-specific pseudonym), and packages may propagate it only if you supply it.

```{r}
library(fluxCore)
set.seed(1)

schema <- default_entity_schema()

# Extend schema without changing Entity/Engine code:
schema$bmi <- list(default = 28, coerce = as.numeric, validate = function(x) length(x) == 1L && (is.na(x) || x > 0))

p <- Entity$new(
  init = list(age = 55, miles_to_work = 10, bmi = 31),
  schema = schema,
  entity_type = "entity",
  time0 = 0
)

prov <- PackageProvider$new()

eng <- Engine$new(
  provider   = prov,
  model_spec = list(name = "default")  # selects default_model_bundle()
)

out <- eng$run(p, max_events = 200, return_observations = TRUE)

tail(out$events, 6)
tail(out$observations, 6)
out$entity$state(c("age", "miles_to_work", "bmi"))
```


## Batch simulation with global parameter draws

```{r}
entities <- lapply(1:4, function(i) {
  Entity$new(
    init = list(age = 45 + i, miles_to_work = 10),
    schema = schema,
    entity_type = "entity",
    time0 = 0
  )
})
names(entities) <- paste0("id", 1:4)

batch <- run_cohort(
  eng,
  entities,
  n_param_draws = 3,
  n_sims = 2,
  max_events = 50,
  backend = "none",
  seed = 123
)

head(batch$index)
```

## Policy / intervention layering

```{r}
# Example: add an intervention flag after time >= 2 (requires schema include tx_on)
schema2 <- schema
schema2$tx_on <- list(default = 0, coerce = as.integer, validate = function(x) x %in% c(0L,1L))

p2 <- Entity$new(
  init = list(age = 55, miles_to_work = 10, tx_on = 0),
  schema = schema2,
  entity_type = "entity",
  time0 = 0
)

baseline <- default_model_bundle()

policy <- list(
  transition = function(entity, event_type, time_next, ctx, baseline_changes = NULL) {
    if (entity$last_time >= 2) return(list(tx_on = 1L))
    NULL
  }
)

bundle2 <- compose_bundles(baseline, policy)

# Build an engine using a custom registry entry
prov2 <- PackageProvider$new(registry = list(composed = function() bundle2))
eng2 <- Engine$new(provider = prov2, model_spec = list(name = "composed"))

out2 <- eng2$run(p2, max_events = 30)
tail(out2$events, 5)
out2$entity$state(c("age","tx_on"))
```
