---
title: "Engine + ModelProvider scaffold"
maturity: "validated"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Engine + ModelProvider scaffold}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette demonstrates the core architecture:

- `Entity` holds mutable state, sparse history, and an event log.
- `Engine` orchestrates the simulation **and materializes models** using a provider + model spec.
- `ModelBundle` supplies the dynamics.
- `ModelProviders` load bundles from different sources.


### Determinism, tie-breaking, and optional entity tags

When multiple processes propose events at the same `time_next`, the engine selects one deterministically by ordering on:

1. earliest `time_next`, then
2. lexicographic `process_id`.

This means `process_id` is not just a label, it is part of the model's deterministic resolution policy. If your model has a preferred priority when ties occur, encode that priority into the `process_id` (for example, a zero-padded numeric prefix).

For downstream linking in a HIPAA-conscious way, a `Entity` may also carry an optional single-column identifier `id` (default `NULL`). This is intended for *de-identified* tags (for example, a study-specific pseudonym), and packages may propagate it only if you supply it.

```{r}
library(fluxCore)
set.seed(1)

schema <- list(
  route_zone = list(
    type = "categorical",
    levels = c("urban", "suburban", "rural"),
    default = "urban",
    coerce = as.character
  )
)

# Extend schema without changing Entity/Engine code:
schema$battery_pct <- list(
  type = "continuous",
  default = 100,
  coerce = as.numeric,
  validate = function(x) length(x) == 1L && (is.na(x) || (x >= 0 && x <= 100))
)
schema$payload_kg <- list(
  type = "continuous",
  default = 0,
  coerce = as.numeric,
  validate = function(x) length(x) == 1L && (is.na(x) || x >= 0)
)

toy_bundle <- list(
  time_spec = time_spec(unit = "hours"),
  propose_events = function(entity, ctx = NULL, process_ids = NULL, current_proposals = NULL) {
    list(dispatch = list(time_next = entity$last_time + 1, event_type = "dispatch"))
  },
  transition = function(entity, event, ctx = NULL) {
    if (!identical(event$event_type, "dispatch")) return(list())
    list(
      payload_kg = max(0, as.numeric(entity$as_list("payload_kg")$payload_kg) + stats::rnorm(1, mean = 1, sd = 0.3)),
      battery_pct = max(0, as.numeric(entity$as_list("battery_pct")$battery_pct) - 4)
    )
  },
  stop = function(entity, event, ctx = NULL) entity$last_time >= 6,
  observe = function(entity, event, ctx = NULL) {
    s <- entity$as_list(c("route_zone", "battery_pct", "payload_kg"))
    data.frame(
      time = entity$last_time,
      event_type = event$event_type,
      route_zone = s$route_zone,
      battery_pct = s$battery_pct,
      payload_kg = s$payload_kg
    )
  }
)

p <- Entity$new(
  init = list(route_zone = "urban", battery_pct = 100, payload_kg = 0),
  schema = schema,
  entity_type = "courier",
  time0 = 0
)

eng <- Engine$new(
  provider = list(load = function(model_spec = NULL, ...) toy_bundle)
)

out <- eng$run(p, max_events = 200, return_observations = TRUE)

tail(out$events, 6)
tail(out$observations, 6)
out$entity$state(c("route_zone", "battery_pct", "payload_kg"))
```


## Batch simulation with global parameter draws

```{r}
entities <- lapply(1:4, function(i) {
  Entity$new(
    init = list(
      route_zone = c("urban", "suburban", "rural", "urban")[i],
      battery_pct = 100 - 5 * i,
      payload_kg = i - 1
    ),
    schema = schema,
    entity_type = "courier",
    time0 = 0
  )
})
names(entities) <- paste0("id", 1:4)

batch <- run_cohort(
  eng,
  entities,
  n_param_draws = 3,
  n_sims = 2,
  max_events = 50,
  backend = "none",
  seed = 123
)

head(batch$index)
```

## Policy / intervention layering

```{r}
# Example: add a priority mode after time >= 2 (requires schema include priority_mode)
schema2 <- schema
schema2$priority_mode <- list(
  type = "categorical",
  levels = c("normal", "surge"),
  default = "normal",
  coerce = as.character
)

p2 <- Entity$new(
  init = list(
    route_zone = "urban",
    battery_pct = 100,
    payload_kg = 0,
    priority_mode = "normal"
  ),
  schema = schema2,
  entity_type = "courier",
  time0 = 0
)

baseline <- toy_bundle

policy <- list(
  transition = function(entity, event, ctx = NULL, baseline_changes = NULL) {
    if (entity$last_time >= 2) return(list(priority_mode = "surge"))
    NULL
  }
)

bundle2 <- compose_bundles(baseline, policy)

eng2 <- Engine$new(provider = list(load = function(model_spec = NULL, ...) bundle2))

out2 <- eng2$run(p2, max_events = 30)
tail(out2$events, 5)
out2$entity$state(c("route_zone", "battery_pct", "priority_mode"))
```
