---
title: "Engine + ModelProvider scaffold"
maturity: "validated"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Engine + ModelProvider scaffold}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette demonstrates the core architecture:

- `Entity` holds mutable state, sparse history, and an event log.
- `Engine` orchestrates the simulation **and materializes models** using a provider + model spec.
- `ModelBundle` supplies the dynamics.
- `ModelProviders` load bundles from different sources.


### Determinism, tie-breaking, and optional entity tags

When multiple processes propose events at the same `time_next`, the engine selects one deterministically by ordering on:

1. earliest `time_next`, then
2. lexicographic `process_id`.

This means `process_id` is not just a label, it is part of the model's deterministic resolution policy. If your model has a preferred priority when ties occur, encode that priority into the `process_id` (for example, a zero-padded numeric prefix).


An `Entity` may also carry an optional single-column identifier `id` (default `NULL`). 


To begin, load the library.
```{r}
library(fluxCore)
set.seed(1)
```

The first task is to define a `schema` for the entity's state variables. The schema specifies the variable names, types, validation rules, and default values. The engine will enforce this schema at runtime, ensuring that state updates are valid and consistent. This separation of concerns allows you to focus on the dynamics in the transition function without worrying about low-level data validation.

The `fluxCore` package provides a default schema specification format, but you can define your own as long as it follows the expected structure. Below, we show how to create a schema using default variable formats (each of which has a default validation strategy), and then how to define a custom schema manually if you need more control.

```{r}
# The schema is a list object with one entry per state variable. Each entry is itself a list with at least a `type` field, and optionally `levels`, `default`, `coerce`, and `validate` fields.
schema <- list(
  route_zone = list(
    type = "categorical",
    levels = c("urban", "suburban", "rural"),
    default = "urban",
    coerce = as.character
  )
)

# Extend schema without changing Entity/Engine code:
schema$battery_pct <- list(
  type = "numeric",
  default = 100,
  coerce = as.numeric
)
schema$payload_kg <- list(
  type = "numeric",
  default = 0,
  coerce = as.numeric
)
```
The above illustrates how to build up a schema incrementally, which is useful when you have a large number of variables or want to define them in different parts of the code. However, if you have a simple model and want to create a schema in one step, the `set_schema` helper function provides a convenient shortcut.

```{r}
quicker_schema <- set_schema(vars = list(
  route_zone           = list(type = "categorical",
                              levels = c("urban", "suburban", "rural")),
  battery_pct          = "percent",
  payload_kg           = list(type = "nonnegative_numeric", max = 20),
  deliveries_completed = "count",
  prob_rain            = "probability"
))
```

This creates a fully validated schema with automatic type-specific defaults. The hybrid `vars` syntax accepts either a type-name string (e.g. `"count"`) or a full list spec (e.g. `list(type = "positive_numeric", max = 20)`) per variable, mixed freely in one call.

You can also extend or replace entries in an existing schema. By default, adding a variable that already exists is an error; pass `overwrite = TRUE` to replace it, or use `remove = ` to drop entries:

```{r}
quicker_schema <- set_schema(
  vars = list(deliveries_completed = list(type = "count", max = 50)),
  schema = quicker_schema,
  overwrite = TRUE
)
```

Each state variable's `type` label gives fluxCore a default validation strategy. Built-in validators are intentionally permissive within each type's semantic — `count` enforces non-negative integer, `probability` enforces [0,1], `percent` enforces [0,100], etc. Use `min` / `max` (numeric types) or a custom `validate` function to tighten further. For a complete list of supported variable types and associated validation rules, see `subrepos/fluxCore/docs/schema_spec.md`.

The `coerce` and `default` fields are optional; `set_schema` will set type-specific defaults if not provided (e.g., `as.numeric` for `numeric` types, `NA_real_` for `numeric` defaults). You can override these if needed.


## Engine, ModelBundle, and ModelProvider

With a schema in hand we can now build the rest of the simulation. fluxCore deliberately separates three concerns:

- **`Entity`** — the *state*. Holds the current values of every schema variable, the sparse history of changes, the event log, and an optional `id`. It validates every update against the schema you just defined. The `Entity` knows nothing about *dynamics*.
- **`ModelBundle`** — the *dynamics*. A plain named list of functions (and a `time_spec`) that together describe how the entity evolves: when events happen, how state changes when an event fires, when the simulation should stop, and what to record. It knows nothing about *where* it came from.
- **`ModelProvider`** — the *source of bundles*. An object with a `$load(model_spec, ...)` method that returns a `ModelBundle`. Providers let you swap how a model is sourced (an in-memory list, a registered package model, a file on disk, a remote service) without touching the engine or the bundle itself. The standard production workflow is to author your model from the `fluxModelTemplate` skeleton and ship it as its own R package; the package then exposes a builder function that a `PackageProvider` can `load()` by name. Inline providers (shown below) are the minimal-ceremony path used for tutorials, scratch work, and tests.

The **`Engine`** ties these together. At construction time it asks its provider to `load()` a bundle; thereafter `engine$run(entity)` executes a deterministic loop: propose candidate events → pick the earliest (with `process_id` lexicographic tie-break) → call `transition()` to compute state changes → `entity$update()` → optionally `observe()` → check `stop()` and `max_time` → refresh proposals as needed.

A minimal `ModelBundle` therefore needs five things:

| Field | Purpose |
|---|---|
| `time_spec` | The clock (units, calendar). Built with `time_spec(unit = ...)`. |
| `propose_events(entity, ctx, process_ids, current_proposals)` | Returns a named list of candidate events, each with `time_next` and `event_type`. The names are the `process_id`s used for tie-breaking. |
| `transition(entity, event, ctx)` | Returns a named list of state changes for the chosen event (or `list()` for a no-op). |
| `stop(entity, event, ctx)` | Returns `TRUE` when the run should terminate. |
| `observe(entity, event, ctx)` | Returns a one-row data frame of the variables you want recorded after each event. Optional — only used when `return_observations = TRUE`. |

Bundles can also expose optional hooks (`init_entity`, `refresh_rules`, `event_catalog`, `params`, `sample_params`); we ignore them here.

### A toy bundle

Below is the smallest interesting bundle: a single process named `dispatch` that fires every hour, adds a noisy parcel of payload, and drains 4% of the battery. It stops at `time = 6` and records the schema-relevant fields after each event.

```{r}
toy_bundle <- list(
  time_spec = time_spec(unit = "hours"),
  propose_events = function(entity, ctx = NULL, process_ids = NULL, current_proposals = NULL) {
    list(dispatch = list(time_next = entity$last_time + 1, event_type = "dispatch"))
  },
  transition = function(entity, event, ctx = NULL) {
    if (!identical(event$event_type, "dispatch")) return(list())
    list(
      payload_kg = max(0, as.numeric(entity$as_list("payload_kg")$payload_kg) + stats::rnorm(1, mean = 1, sd = 0.3)),
      battery_pct = max(0, as.numeric(entity$as_list("battery_pct")$battery_pct) - 4)
    )
  },
  stop = function(entity, event, ctx = NULL) entity$last_time >= 6,
  observe = function(entity, event, ctx = NULL) {
    s <- entity$as_list(c("route_zone", "battery_pct", "payload_kg", "deliveries_completed", "prob_rain"))
    data.frame(
      time = entity$last_time,
      event_type = event$event_type,
      route_zone = s$route_zone,
      battery_pct = s$battery_pct,
      payload_kg = s$payload_kg,
      deliveries_completed = s$deliveries_completed,
      prob_rain = s$prob_rain
    )
  }
)
```

Notice that nothing in `toy_bundle` knows the schema directly — the bundle reads variables through `entity$as_list()` and writes them back via the named list returned by `transition()`. The engine will run those returned values through schema validation before they are committed to the entity.

### Constructing the Entity

The `Entity` is initialized with a starting state (one value per schema variable) and the schema we built above. The engine will reject any subsequent update that violates that schema.

```{r}
p <- Entity$new(
  init = list(route_zone = "urban", battery_pct = 100, payload_kg = 0, deliveries_completed = 0L, prob_rain = 0.5),
  schema = quicker_schema,
  entity_type = "courier",
  time0 = 0
)
```

### Wiring it together: ModelProvider and Engine

The `Engine` does not accept a bundle directly — it accepts a `ModelProvider` and asks the provider to `load()` a bundle for it. This indirection is what lets the same engine code drive a quick in-memory bundle (here) and a full package-backed model (in real projects) with no changes.

A `ModelProvider` is just any object with a `$load(model_spec, ...)` method that returns a valid `ModelBundle`. The cheapest one we can write is a one-element list whose `load` function ignores its arguments and returns `toy_bundle`:

```{r}
toy_provider <- list(load = function(model_spec = NULL, ...) toy_bundle)
```

That is a complete, valid `ModelProvider`. It satisfies the same contract as `PackageProvider$new(registry = list(courier = build_courier_bundle))` would in a packaged model — the engine cannot tell the difference.

With provider in hand, building and running the engine is mechanical:

```{r}
eng <- Engine$new(provider = toy_provider)

out <- eng$run(p, max_events = 200, return_observations = TRUE)

tail(out$events, 6)
tail(out$observations, 6)
out$entity$state(c("route_zone", "battery_pct", "payload_kg"))
```

`out$events` is the engine's authoritative event log; `out$observations` is whatever your `observe()` hook accumulated; and `out$entity` is the same `p` from above, now mutated with the post-run state.


## Batch simulation with global parameter draws

```{r}
entities <- lapply(1:4, function(i) {
  Entity$new(
    init = list(
      route_zone = c("urban", "suburban", "rural", "urban")[i],
      battery_pct = 100 - 5 * i,
      payload_kg = i - 1,
      deliveries_completed = 0L,
      prob_rain = 0.5
    ),
    schema = quicker_schema,
    entity_type = "courier",
    time0 = 0
  )
})
names(entities) <- paste0("id", 1:4)

batch <- run_cohort(
  eng,
  entities,
  n_param_draws = 3,
  n_sims = 2,
  max_events = 50,
  backend = "none",
  seed = 123
)

head(batch$index)
```

## Policy / intervention layering

```{r}
# Example: add a priority mode after time >= 2 (requires schema include priority_mode)
schema2 <- quicker_schema
schema2$priority_mode <- list(
  type = "categorical",
  levels = c("normal", "surge"),
  default = "normal",
  coerce = as.character
)

p2 <- Entity$new(
  init = list(
    route_zone = "urban",
    battery_pct = 100,
    payload_kg = 0,
    deliveries_completed = 0L,
    prob_rain = 0.5,
    priority_mode = "normal"
  ),
  schema = schema2,
  entity_type = "courier",
  time0 = 0
)

baseline <- toy_bundle

policy <- list(
  transition = function(entity, event, ctx = NULL, baseline_changes = NULL) {
    if (entity$last_time >= 2) return(list(priority_mode = "surge"))
    NULL
  }
)

bundle2 <- compose_bundles(baseline, policy)

eng2 <- Engine$new(provider = list(load = function(model_spec = NULL, ...) bundle2))

out2 <- eng2$run(p2, max_events = 30)
tail(out2$events, 5)
out2$entity$state(c("route_zone", "battery_pct", "priority_mode"))
```
