---
title: "Engine and ModelBundle scaffold"
maturity: "validated"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Engine and ModelBundle scaffold}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
suppressPackageStartupMessages(library(fluxCore))
```

This vignette demonstrates the core architecture:

- `Entity` holds mutable state, sparse history, and an event log.
- `ModelBundle` supplies the dynamics (when events happen, how state changes, when to stop, what to record).
- `Engine` orchestrates the simulation by repeatedly proposing, picking, applying, and observing events.

If you haven't read [00_start_here.md](00_start_here.md) yet, do that first — it covers the conceptual framework (entities, processes, decisions, the single-entity → multi-entity spectrum) that this tutorial puts into code.

This tutorial uses the simpler direct `bundle =` constructor and defers `ModelProvider` and package-based model loading to a later tutorial.

### Determinism, tie-breaking, and optional entity tags

When multiple processes propose events at the same `time_next`, the engine selects one deterministically by ordering on:

1. earliest `time_next`, then
2. lexicographic `process_id`.

This means `process_id` is not just a label, it is part of the model's deterministic resolution policy. If your model has a preferred priority when ties occur, encode that priority into the `process_id` (for example, a zero-padded numeric prefix).


An `Entity` may also carry an optional single-column identifier `id` (default `NULL`). 


To begin, load the library.
```{r}
library(fluxCore)
set.seed(1)
```

The first task is to define a `schema` for the entity's state variables. The schema specifies the variable names, types, validation rules, and default values. The engine will enforce this schema at runtime, ensuring that state updates are valid and consistent. This separation of concerns allows you to focus on the dynamics in the transition function without worrying about low-level data validation.

The `fluxCore` package provides a default schema specification format, but you can define your own as long as it follows the expected structure. Below, we show how to create a schema using default variable formats (each of which has a default validation strategy), and then how to define a custom schema manually if you need more control.

```{r}
# The schema is a list object with one entry per state variable. Each entry is itself a list with at least a `type` field, and optionally `levels`, `default`, `coerce`, and `validate` fields.
schema <- list(
  route_zone = list(
    type = "categorical",
    levels = c("urban", "suburban", "rural"),
    default = "urban",
    coerce = as.character
  )
)

# Extend schema without changing Entity/Engine code:
schema$battery_pct <- list(
  type = "numeric",
  default = 100,
  coerce = as.numeric
)
schema$payload_kg <- list(
  type = "numeric",
  default = 0,
  coerce = as.numeric
)
```
The above illustrates how to build up a schema incrementally, which is useful when you have a large number of variables or want to define them in different parts of the code. However, if you have a simple model and want to create a schema in one step, the `set_schema` helper function provides a convenient shortcut.

```{r}
quicker_schema <- set_schema(vars = list(
  route_zone           = list(type = "categorical",
                              levels = c("urban", "suburban", "rural")),
  battery_pct          = "percent",
  payload_kg           = list(type = "nonnegative_numeric", max = 20),
  deliveries_completed = "count",
  prob_rain            = "probability"
))
```

This creates a fully validated schema with automatic type-specific defaults. The hybrid `vars` syntax accepts either a type-name string (e.g. `"count"`) or a full list spec (e.g. `list(type = "positive_numeric", max = 20)`) per variable, mixed freely in one call.

You can also extend or replace entries in an existing schema. By default, adding a variable that already exists is an error; pass `overwrite = TRUE` to replace it, or use `remove = ` to drop entries:

```{r}
quicker_schema <- set_schema(
  vars = list(deliveries_completed = list(type = "count", max = 50)),
  schema = quicker_schema,
  overwrite = TRUE
)
```

Each state variable's `type` label gives fluxCore a default validation strategy. Built-in validators are intentionally permissive within each type's semantic — `count` enforces non-negative integer, `probability` enforces [0,1], `percent` enforces [0,100], etc. Use `min` / `max` (numeric types) or a custom `validate` function to tighten further. For a complete list of supported variable types and associated validation rules, see `subrepos/fluxCore/docs/schema_spec.md`.

The `coerce` and `default` fields are optional; `set_schema` will set type-specific defaults if not provided (e.g., `as.numeric` for `numeric` types, `NA_real_` for `numeric` defaults). You can override these if needed.


## Engine and ModelBundle

With a schema in hand we can now build the rest of the simulation. fluxCore deliberately separates two concerns at this level:

- **`Entity`** — the *state*. Holds the current values of every schema variable, the sparse history of changes, the event log, and an optional `id`. It validates every update against the schema you just defined. The `Entity` knows nothing about *dynamics*.
- **`ModelBundle`** — the *dynamics*. A plain named list of functions (and a `time_spec`) that together describe how the entity evolves: when events happen, how state changes when an event fires, when the simulation should stop, and what to record.

The **`Engine`** ties these together. You hand it a bundle; it then runs a deterministic loop on a given `Entity`: propose candidate events → pick the earliest (with `process_id` lexicographic tie-break) → call `transition()` to compute state changes → `entity$update()` → optionally `observe()` → check `stop()` and `max_time` → refresh proposals as needed.

A minimal `ModelBundle` therefore needs five things:

| Field | Purpose |
|---|---|
| `time_spec` | The clock (units, calendar). Built with `time_spec(unit = ...)`. |
| `propose_events(entity, ctx, process_ids, current_proposals)` | Returns a named list of candidate events, each with `time_next` and `event_type`. The names are the `process_id`s used for tie-breaking. |
| `transition(entity, event, ctx)` | Returns a named list of state changes for the chosen event (or `list()` for a no-op). |
| `stop(entity, event, ctx)` | Returns `TRUE` when the run should terminate. |
| `observe(entity, event, ctx)` | Returns a one-row data frame of the variables you want recorded after each event. Optional — only used when `return_observations = TRUE`. |

Bundles can also expose optional hooks (`init_entity`, `refresh_rules`, `event_catalog`, `params`, `sample_params`); we ignore them here.

> *Aside on `ModelProvider`.* In production you typically don't author a bundle inline — you build it from the `fluxModelTemplate` skeleton, ship it as its own R package, and register it through a `ModelProvider` (e.g. `PackageProvider$new(registry = list(my_model = build_my_model))`). The `Engine` then asks the provider to `load()` a bundle by `model_spec`, which makes runs reproducible and swappable. We use the simpler direct path (`Engine$new(bundle = ...)`) for the rest of this tutorial; a forthcoming tutorial covers providers in full.

### A toy bundle

Below is the smallest interesting bundle: a single process named `dispatch` that fires every hour, adds a noisy parcel of payload, and drains 4% of the battery. It stops at `time = 6` and records the schema-relevant fields after each event.

```{r}
toy_bundle <- list(
  time_spec = time_spec(unit = "hours"),
  propose_events = function(entity, ctx = NULL, process_ids = NULL, current_proposals = NULL) {
    list(dispatch = list(time_next = entity$last_time + 1, event_type = "dispatch"))
  },
  transition = function(entity, event, ctx = NULL) {
    if (!identical(event$event_type, "dispatch")) return(list())
    list(
      payload_kg = max(0, as.numeric(entity$as_list("payload_kg")$payload_kg) + stats::rnorm(1, mean = 1, sd = 0.3)),
      battery_pct = max(0, as.numeric(entity$as_list("battery_pct")$battery_pct) - 4)
    )
  },
  stop = function(entity, event, ctx = NULL) entity$last_time >= 6,
  observe = function(entity, event, ctx = NULL) {
    s <- entity$as_list(c("route_zone", "battery_pct", "payload_kg", "deliveries_completed", "prob_rain"))
    data.frame(
      time = entity$last_time,
      event_type = event$event_type,
      route_zone = s$route_zone,
      battery_pct = s$battery_pct,
      payload_kg = s$payload_kg,
      deliveries_completed = s$deliveries_completed,
      prob_rain = s$prob_rain
    )
  }
)
```

Notice that nothing in `toy_bundle` knows the schema directly — the bundle reads variables through `entity$as_list()` and writes them back via the named list returned by `transition()`. The engine will run those returned values through schema validation before they are committed to the entity.

### Constructing the Entity

The `Entity` is initialized with a starting state (one value per schema variable) and the schema we built above. The engine will reject any subsequent update that violates that schema.

```{r}
p <- Entity$new(
  init = list(route_zone = "urban", battery_pct = 100, payload_kg = 0, deliveries_completed = 0L, prob_rain = 0.5),
  schema = quicker_schema,
  entity_type = "courier",
  time0 = 0
)
```

### Constructing the Engine and running

With a bundle in hand, building and running the engine is one line. Pass the bundle directly via `Engine$new(bundle = ...)`; the engine validates the bundle and is then ready to run.

```{r}
eng <- Engine$new(bundle = toy_bundle)

out <- eng$run(p, max_events = 200, return_observations = TRUE)

tail(out$events, 6)
tail(out$observations, 6)
out$entity$state(c("route_zone", "battery_pct", "payload_kg"))
```

`out$events` is the engine's authoritative event log; `out$observations` is whatever your `observe()` hook accumulated; and `out$entity` is the same `p` from above, now mutated with the post-run state.


## Variable blocks: grouped state updates

Some state variables naturally evolve **as a unit**. In our delivery domain, *weather* is a good example: when the weather refreshes, both the chance of rain and the wind speed change together, on the same clock, from the same draw. fluxCore lets you tag those variables as members of a named **block**, then update the whole block in one call from `transition()`.

To demonstrate, we'll extend the schema with a new variable `wind_kph` and tag both `wind_kph` and the existing `prob_rain` as members of a `weather` block. Because `prob_rain` already exists in `quicker_schema`, we use `overwrite = TRUE` to replace its spec with one that carries the `blocks` field:

```{r}
quicker_schema_w_weather <- set_schema(
  vars = list(
    prob_rain = list(type = "probability", blocks = "weather"),
    wind_kph  = list(type = "nonnegative_numeric", max = 100, blocks = "weather")
  ),
  schema = quicker_schema,
  overwrite = TRUE
)

schema_blocks(quicker_schema_w_weather)
block_vars(quicker_schema_w_weather, "weather")
```

A variable can belong to **multiple** blocks; the `blocks` field is a character vector and membership is many-to-many. Block membership is purely metadata on the schema — the engine itself does not interpret blocks. They are a contract that helpers like `update_block()` and `combine_updates()` use to validate that a multivariate update is well-formed.

Now we'll build a bundle with two competing processes:

- a `dispatch` process (as before, with a random inter-arrival time so it doesn't collide with weather), and
- a `weather` process that fires on its own clock and updates the entire `weather` block at once via `update_block()`.

```{r}
weather_aware_bundle <- list(
  time_spec = time_spec(unit = "hours"),

  propose_events = function(entity, ctx = NULL, process_ids = NULL, current_proposals = NULL) {
    list(
      dispatch = list(
        time_next  = entity$last_time + stats::rexp(1, rate = 1.0),
        event_type = "dispatch"
      ),
      weather = list(
        time_next  = entity$last_time + stats::rexp(1, rate = 0.5),
        event_type = "weather_refresh"
      )
    )
  },

  transition = function(entity, event, ctx = NULL) {
    if (identical(event$event_type, "dispatch")) {
      return(list(
        payload_kg  = max(0, as.numeric(entity$as_list("payload_kg")$payload_kg) +
                              stats::rnorm(1, mean = 1, sd = 0.3)),
        battery_pct = max(0, as.numeric(entity$as_list("battery_pct")$battery_pct) - 4)
      ))
    }
    if (identical(event$event_type, "weather_refresh")) {
      # Single call updates *both* weather variables; require_all = TRUE (default)
      # forces the caller to supply every block member, which catches typos and
      # keeps multivariate state consistent.
      return(update_block(entity, "weather", list(
        prob_rain = stats::runif(1, 0, 1),
        wind_kph  = max(0, stats::rnorm(1, mean = 15, sd = 5))
      )))
    }
    list()
  },

  stop = function(entity, event, ctx = NULL) entity$last_time >= 6,

  observe = function(entity, event, ctx = NULL) {
    s <- entity$as_list(c("battery_pct", "payload_kg", "prob_rain", "wind_kph"))
    data.frame(
      time        = entity$last_time,
      event_type  = event$event_type,
      battery_pct = s$battery_pct,
      payload_kg  = round(s$payload_kg, 2),
      prob_rain   = round(s$prob_rain, 2),
      wind_kph    = round(s$wind_kph, 1)
    )
  }
)
```

The key line is `update_block(entity, "weather", list(prob_rain = ..., wind_kph = ...))`. It does three things: (1) checks that `weather` is a declared block in the entity's schema, (2) verifies that the supplied names cover every block member (`require_all = TRUE` is the default), and (3) returns a schema-validated, schema-ordered named list ready to be returned from `transition()`. If you forgot one member, mistyped a name, or supplied a value that violates its variable's type, you get a clear error here rather than a silent half-applied update.

Running it:

```{r}
set.seed(2)
p_w <- Entity$new(
  init = list(
    route_zone           = "urban",
    battery_pct          = 100,
    payload_kg           = 0,
    deliveries_completed = 0L,
    prob_rain            = 0.5,
    wind_kph             = 12
  ),
  schema = quicker_schema_w_weather,
  entity_type = "courier",
  time0 = 0
)

eng_w <- Engine$new(bundle = weather_aware_bundle)
out_w <- eng_w$run(p_w, max_events = 200, return_observations = TRUE)
tail(out_w$observations, 10)
```

Notice how `prob_rain` and `wind_kph` only ever change on rows whose `event_type` is `weather_refresh`, and they always change *together*. That's the contract the block buys you.


## Batch simulation with global parameter draws

```{r}
entities <- lapply(1:4, function(i) {
  Entity$new(
    init = list(
      route_zone = c("urban", "suburban", "rural", "urban")[i],
      battery_pct = 100 - 5 * i,
      payload_kg = i - 1,
      deliveries_completed = 0L,
      prob_rain = 0.5
    ),
    schema = quicker_schema,
    entity_type = "courier",
    time0 = 0
  )
})
names(entities) <- paste0("id", 1:4)

batch <- run_cohort(
  eng,
  entities,
  n_param_draws = 3,
  n_sims = 2,
  max_events = 50,
  backend = "none",
  seed = 123
)

head(batch$index)
```

## Decision points, actions, and trajectory logging

Everything so far has been a *pure generative process*: events happen, state
evolves, the engine stops when `stop()` returns `TRUE`. There is no external
agent choosing what to do.

Many real systems involve **decisions** — moments where a policy (a clinician, a
dispatcher, a learned algorithm) is consulted and may inject an intervention into
the timeline. fluxCore makes these first-class through three structures that work
together:

| Structure | What it is |
|---|---|
| `DecisionPoint` | *Where* in the simulation a decision can occur. Declared in the schema. Triggers on specified event types. |
| `ActionEvent` | The intervention itself. Proposed by the policy at a decision point; enters the normal event timeline. |
| `TrajectoryRecord` | The audit log emitted at each decision point: what was observed, what action was proposed, what was realized. |

The key conceptual move is that **actions are events**. A policy does not mutate
state directly — it proposes an `ActionEvent` with a `time_next` and an
`action_type`. That action enters the same proposal queue as any other event and
is realized by the normal `transition()` / `stop()` path. This means:

- Actions obey the same deterministic ordering as everything else.
- Actions can be ignored, arbitrated against, or superseded like any other event.
- The full trajectory — what the entity saw, what the policy proposed, what
  actually happened — is captured in `TrajectoryRecord` objects for later audit,
  comparison, or RL reward computation.

### Declaring a decision point

Decision points are declared on the schema, not inferred at runtime. This makes
the policy interface explicit: any schema without `$decision_points` runs as a
pure generative model regardless of what policy you pass.

For our delivery example, a natural decision point is *after each dispatch*: once
a delivery is completed, should the drone surge to priority mode or continue
normally?

```{r}
# Variables schema for the decision-point example
dp_vars <- set_schema(
  vars = list(
    route_zone           = list(type = "categorical",
                                levels = c("urban", "suburban", "rural")),
    battery_pct          = "percent",
    payload_kg           = list(type = "nonnegative_numeric", max = 20),
    deliveries_completed = "count",
    prob_rain            = "probability",
    priority_mode        = list(type = "categorical",
                                levels = c("normal", "surge"),
                                default = "normal")
  )
)

# Declare a decision point: fires after every "dispatch" event.
# The policy may propose the action "surge" or "stand_down".
dp_post_dispatch <- DecisionPoint(
  id              = "post_dispatch",
  trigger         = "dispatch",
  allowed_actions = c("surge", "stand_down"),
  label           = "After each delivery: surge or stand down?"
)

# Full schema for load_model(): wraps variables, time_spec, and decision_points
schema_dp <- list(
  variables       = dp_vars,
  time_spec       = time_spec(unit = "hours"),
  decision_points = list(dp_post_dispatch)
)
```

`trigger` is the event type(s) that fire the decision point. The engine checks
this after applying `transition()` — so the entity's state already reflects the
just-completed event when the policy is consulted. `allowed_actions` constrains
what the policy is permitted to propose; actions outside this set are ignored with
a warning.

You can also supply an `observation_fn` — a function of `entity` that extracts a
compact, policy-relevant view of state. If you omit it, the policy receives the
full entity.

### Writing a policy

A policy is a list with a `propose_action` method:

```
propose_action(decision_point, entity, sim_ctx, param_ctx) -> ActionEvent | NULL
```

It returns an `ActionEvent` if it wants to intervene, or `NULL` for no
intervention. Here is a simple rule-based policy: surge if battery is below 60%,
stand down otherwise.

```{r}
battery_policy <- list(
  propose_action = function(decision_point, entity, sim_ctx, param_ctx) {
    battery <- entity$current$battery_pct

    action <- if (!is.null(battery) && battery < 60) "surge" else "stand_down"

    ActionEvent(
      action_type       = action,
      time_next         = entity$last_time + 0.05,   # realized 3 minutes later
      decision_point_id = decision_point$id,
      metadata          = list(battery_at_decision = battery)
    )
  }
)
```

`time_next` on an `ActionEvent` is the time at which the action is realized in
the simulation — here, 0.05 hours (3 minutes) after the triggering dispatch. The
action then flows through `transition()` just like any other event.

### Assembling with load_model() and enabling trajectory logging

When decision points are present, the recommended assembly path is `load_model()`
rather than `Engine$new()`. It validates that the schema, bundle, policy, and
(optionally) trajectory logger are all mutually consistent before any run begins.

`trajectory` controls what gets captured at each decision point:

- `list(detail = "none")` — records that the decision point fired, but no state snapshots.
- `list(detail = "summary")` — before/after snapshots via `state_summary_default()`.
- `list(detail = "full")` — complete `entity$current` captured before and after.

```{r}
# Bundle that handles the "surge" and "stand_down" action types in transition()
dp_bundle <- list(
  time_spec = time_spec(unit = "hours"),

  propose_events = function(entity, sim_ctx = NULL, param_ctx = NULL,
                            process_ids = NULL, current_proposals = NULL) {
    list(dispatch = list(
      time_next  = entity$last_time + stats::rexp(1, rate = 1.0),
      event_type = "dispatch"
    ))
  },

  transition = function(entity, event, sim_ctx = NULL, param_ctx = NULL) {
    if (identical(event$event_type, "dispatch")) {
      return(list(
        # Pick up and deliver a package; decrease battery
        payload_kg           = 0,
        battery_pct          = max(0, as.numeric(entity$current$battery_pct) - 4),
        deliveries_completed = as.integer(entity$current$deliveries_completed) + 1L
      ))
    }
    if (identical(event$event_type, "surge"))      return(list(priority_mode = "surge"))
    if (identical(event$event_type, "stand_down")) return(list(priority_mode = "normal"))
    list()
  },

  stop = function(entity, event, sim_ctx = NULL, param_ctx = NULL) {
    as.numeric(entity$current$battery_pct) < 10
  },

  observe = function(entity, event, sim_ctx = NULL, param_ctx = NULL) {
    data.frame(
      time                 = entity$last_time,
      event_type           = event$event_type,
      battery_pct          = entity$current$battery_pct,
      deliveries_completed = entity$current$deliveries_completed,
      priority_mode        = entity$current$priority_mode
    )
  }
)

# Assemble the engine with policy and trajectory logging enabled
eng_dp <- load_model(
  schema     = schema_dp,
  bundle     = dp_bundle,
  policy     = battery_policy,
  trajectory = list(detail = "summary")
)
```

### Running and inspecting trajectory records

```{r}
set.seed(7)
p_dp <- Entity$new(
  init = list(
    route_zone           = "urban",
    battery_pct          = 100,
    payload_kg           = 0,
    deliveries_completed = 0L,
    prob_rain            = 0.3,
    priority_mode        = "normal"
  ),
  schema      = schema_dp$variables,
  entity_type = "courier",
  time0       = 0
)

out_dp <- eng_dp$run(p_dp, max_events = 200, return_observations = TRUE)
```

`out_dp$trajectory_records` is a list of named lists — one per decision point
firing. Each record answers four questions:

1. **When did the decision point fire?** (`t`, `decision_point_id`)
2. **What did the policy observe?** (`observation` — the state snapshot presented to the policy)
3. **What did the policy propose?** (`proposed_actions`)
4. **What actually happened?** (`realized_event`, `state_before`, `state_after`)

```{r}
# How many decision points fired?
length(out_dp$trajectory_records)

# Inspect the first record
tr1 <- out_dp$trajectory_records[[1]]
cat("Time:          ", tr1$t, "\n")
cat("Decision point:", tr1$decision_point_id, "\n")
cat("Action taken:  ", tr1$realized_event$event_type, "\n")
cat("Battery before:", tr1$state_before$battery_pct, "\n")
cat("Battery after: ", tr1$state_after$battery_pct, "\n")
```

```{r}
# Build a summary table across all trajectory records
tr_df <- do.call(rbind, lapply(out_dp$trajectory_records, function(tr) {
  data.frame(
    t              = tr$t,
    action_taken   = tr$realized_event$event_type,
    battery_before = tr$state_before$battery_pct,
    deliveries     = tr$state_before$deliveries_completed,
    stringsAsFactors = FALSE
  )
}))

head(tr_df, 10)
```

Notice the pattern: the policy switches to "surge" as the battery drops below
60%, then back to "stand_down" once an action is re-evaluated. Because
`TrajectoryRecord` captures state before and after each decision, you can
reconstruct exactly what the policy saw and why it acted — which is the foundation
for policy comparison, counterfactual analysis, and RL reward computation in
`fluxSimulate`.

### What trajectory records enable downstream

`TrajectoryRecord` is not just an audit log. It is the structured surface that the
rest of the ecosystem builds on:

- **Policy comparison** (`fluxSimulate`): run the same entity under two policies
  with the same seed; diff the `trajectory_records` to see where they diverge.
- **Counterfactual analysis**: fix the seed and parameter draw; vary only the
  policy; compare outcomes.
- **RL training** (`fluxSimulate`): the `(observation, action, reward,
  next_state)` tuple for each record becomes a training transition. The reward
  function is defined externally and applied to the record.
- **Audit and explainability**: for any individual run, you can reconstruct the
  full decision history and ask "why did the simulation do that?"
