---
title: "Cohort simulation and forecast"
maturity: "draft"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort simulation and forecast}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
suppressPackageStartupMessages({
  library(fluxCore)
  library(fluxForecast)
})
```

This tutorial shows how to go from a single-entity simulation (Tutorial 01) to
**cohort-level forecasting** across many independent entities. This is not an
agent-based model (ABM): couriers do not interact with each other here. A
future ABM tutorial will cover interacting entities and shared environments.

By the end you will be able to:
- build a heterogeneous cohort of couriers,
- run the cohort through an 8-hour shift,
- generate probabilistic forecasts of delivery events and battery state,
- summarize aggregate fleet outcomes (for example, total deliveries over the shift).

We use the urban food delivery model throughout. If you haven't worked through
[01_core_engine_scaffold.md](01_core_engine_scaffold.md), start there — it
covers the `Entity`, `ModelBundle`, and `Engine` concepts that this tutorial
builds on.

## Load the model

The delivery model lives in a plain R script. Sourcing it gives you
`delivery_bundle()`, `delivery_schema()`, and the individual callback functions.
No package install required.

```{r}
source("tutorials/model/urban_delivery.R")
```

## Build a heterogeneous cohort

A **cohort** is just a named list of `Entity` objects. Each agent starts its
shift with a different battery level, home zone, and starting state — exactly
the kind of heterogeneity you see in a real fleet.

This is also where you can plug in a synthetic population generator if you have
one. For example, you might sample initial state from external population
distributions, then instantiate one `Entity` per sampled courier profile.

```{r}
set.seed(2026)

n_couriers <- 20
shared_schema <- delivery_schema()

couriers <- lapply(seq_len(n_couriers), function(i) {
  Entity$new(
    id   = paste0("driver_", sprintf("%02d", i)),
    init = list(
      battery_pct   = runif(1, min = 50, max = 100),
      route_zone    = sample(c("urban", "suburban", "rural"), 1,
                             prob = c(0.55, 0.30, 0.15)),
      payload_kg    = 0,
      dispatch_mode = "idle"
    ),
    schema      = shared_schema,
    entity_type = "courier",
    time0       = 0
  )
})
names(couriers) <- sapply(couriers, function(e) e$id)
```

Quick sanity check — the fleet's starting battery distribution:

```{r}
batteries <- sapply(couriers, function(e) e$current$battery_pct)
battery_summary <- as.data.frame(as.list(summary(batteries)))
knitr::kable(battery_summary, digits = 2)
```

## Single-agent run

Before running the full cohort, let's step through one courier to see the shape
of the output. The `Engine` is constructed once from the bundle and then reused
for every courier.

```{r}
eng <- Engine$new(bundle = delivery_bundle())

out_single <- eng$run(couriers[[1]], max_events = 500, return_observations = TRUE)
```

The result contains:
- `$events` — the authoritative event log (one row per realized event)
- `$observations` — whatever the `observe()` hook emitted
- `$entity` — the same entity object, now mutated to post-run state

```{r}
nrow(out_single$events)
knitr::kable(tail(out_single$observations, 5), digits = 3)
out_single$entity$state(c("battery_pct", "dispatch_mode"))
```

The final event should be `end_shift` — the model's terminal event. The battery
will be lower than it started, and the courier may have completed several
deliveries during the shift.

## Cohort simulation

`run_cohort()` runs the engine over every entity in the list, optionally with
multiple parameter draws (for uncertainty quantification) and multiple simulation
replicates per draw.

```{r}
cohort_result <- run_cohort(
  eng,
  entities      = couriers,
  n_param_draws = 1,
  n_sims        = 50,
  max_events    = 500,
  seed          = 42
)
```

Here:
- `n_param_draws = 1` uses one parameter set (no parameter uncertainty yet),
- `n_sims = 50` runs 50 stochastic simulations per courier,
- `max_events = 500` caps events per run for safety,
- `seed = 42` makes the run reproducible.

The result is an indexed list of run outputs. `cohort_result$index` tells you
which courier/parameter-draw/simulation each slot corresponds to.

- `sim_id` is the simulation replicate number within a courier and parameter draw.
- `run_id` is the unique row key for each realized run.

```{r}
knitr::kable(head(cohort_result$index))

# Example aggregate: total completed deliveries per run over the shift
delivery_counts <- sapply(cohort_result$runs, function(r) {
  sum(r$events$event_type == "delivery_completed")
})

delivery_summary <- data.frame(
  mean = mean(delivery_counts),
  p25 = as.numeric(stats::quantile(delivery_counts, 0.25)),
  p50 = as.numeric(stats::quantile(delivery_counts, 0.50)),
  p75 = as.numeric(stats::quantile(delivery_counts, 0.75)),
  max = max(delivery_counts)
)

knitr::kable(delivery_summary, digits = 2,
             caption = "Total deliveries per run over 8 hours")
```

## Forecasting

`fluxForecast` is one major part of the ecosystem. It takes raw simulation
output and turns it into queryable probabilistic summaries of the future.

### `forecast()` — the entry point

`forecast()` wraps the cohort run into a forecast object that downstream
functions can query. It needs the engine, the entities, and the evaluation
times (the "horizon grid" at which we want predictions).

```{r}
times <- seq(0, 8, by = 1)  # include baseline hour 0 for explicit start_time

fc <- forecast(
  engine   = eng,
  entities = couriers,
  times    = times,
  S        = 80,       # forecast simulation draws per courier
  vars     = "battery_pct",
  seed     = 42
)
```

`S` controls the number of forecast draws generated inside `forecast()`. It is
separate from `n_sims` in `run_cohort()`, which controlled the earlier example
run. They can be the same, but they do not need to be.

If you see the warning about missing `alive`, that is expected for this model.
`fluxForecast` then derives lifecycle status from terminal events, which keeps
time-to-event summaries such as `event_prob()` well-defined.

### `event_prob()` — probability of a delivery event

"What is the probability that each courier completes at least one delivery by
hour *t*?"

```{r}
ep <- event_prob(fc, event = "delivery_completed", times = times)
knitr::kable(head(ep$result), digits = 3)
```

`ep` is a `flux_event_prob` object. The per-time summary table is in
`ep$result`.

If you previously saw `spec$start_time = 1`, that happened because the forecast
time grid started at hour 1. Here we include hour 0 in `times`, so
`start_time` defaults to 0.

By default this is an aggregate cohort risk curve. It starts near 0 at hour 0
and approaches 1 by hour 8 (most couriers complete at least one delivery).

Agents starting with low battery tend to have lower delivery probability early
on:

```{r}
# Compare the 5 lowest-battery couriers vs the 5 highest
low_bat  <- names(sort(batteries))[1:5]
high_bat <- names(sort(batteries, decreasing = TRUE))[1:5]

id_map <- unique(fc$run_index[, c("entity_id", "entity_tag")])
delivery_col <- match("delivery_completed", fc$event_types)
delivery_by_h4 <- fc$first_event_time[, delivery_col] <= 4

courier_risk_h4 <- aggregate(
  delivery_by_h4,
  by = list(entity_id = fc$run_index$entity_id,
            entity_tag = fc$run_index$entity_tag),
  FUN = mean
)
names(courier_risk_h4)[names(courier_risk_h4) == "x"] <- "event_prob_h4"

ep_low  <- courier_risk_h4[courier_risk_h4$entity_tag %in% low_bat, ]
ep_high <- courier_risk_h4[courier_risk_h4$entity_tag %in% high_bat, ]

cat("Mean P(delivery by hour 4) — low battery group: ",
  round(mean(ep_low$event_prob_h4), 3), "\n")
cat("Mean P(delivery by hour 4) — high battery group:",
  round(mean(ep_high$event_prob_h4), 3), "\n")
```

### `state_summary()` — distribution of a state variable

"What does the battery distribution look like across the fleet at each hour?"

```{r}
ss <- state_summary(fc, vars = "battery_pct", times = times, by = "entity")
ss_battery <- merge(ss$battery_pct, id_map, by = "entity_id", all.x = TRUE)
knitr::kable(head(ss_battery), digits = 3)
```

This gives you quantiles (or other summary statistics) of `battery_pct` at each
evaluation time. As the shift progresses, the distribution shifts left — agents
drain their batteries at different rates depending on their dispatch/delivery
intensity.

### `draws()` — inspect raw trajectories

For a single agent, you can pull the underlying simulation draws to see the
stochastic spread:

```{r}
dr <- draws(fc, var = "battery_pct", times = times, start_time = 0)
dr <- merge(dr, fc$run_index[, c("run_id", "entity_tag")], by = "run_id")

# Filter to one courier
dr_one <- dr[dr$entity_tag == "driver_01", ]
knitr::kable(head(dr_one, 10), digits = 3)
```

Each row is one draw × one time point. You get the full distribution rather than
just a summary — useful for checking whether the forecast is well-behaved or if
there are pathological outliers.

## Varying model parameters

One of the design goals of the bundle architecture is that you can swap
parameters without changing any other code. Let's compare the default dispatch
rate against a slower fleet:

```{r}
# Default: dispatch_rate_base = 0.7
eng_slow <- Engine$new(bundle = delivery_bundle(
  params = list(dispatch_rate_base = 0.3)
))

fc_slow <- forecast(
  engine   = eng_slow,
  entities = couriers,
  times    = times,
  S        = 80,
  vars     = "battery_pct",
  seed     = 42
)

ep_slow <- event_prob(fc_slow, event = "delivery_completed", times = times)
```

Compare mean fleet-wide delivery probability at hour 4:

```{r}
cat("Default dispatch rate — P(delivery by hour 4):",
  round(ep$result$event_prob[ep$result$time == 4], 3), "\n")
cat("Slow dispatch rate   — P(delivery by hour 4):",
  round(ep_slow$result$event_prob[ep_slow$result$time == 4], 3), "\n")
```

The slower dispatch rate produces a flatter event probability curve — agents
receive fewer assignments, so fewer deliveries are completed by any given hour.
This is exactly the kind of "what if" scenario that fleet operators care about:
if demand drops (lower dispatch rate), how does delivery throughput change?

## Summary

| Concept | What you learned |
|---------|-----------------|
| `Entity` cohort | A named list of couriers (independent entities) with heterogeneous starting state |
| `run_cohort()` | Batch simulation with parameter draws and replicates |
| `forecast()` | Wraps cohort output into a queryable forecast object |
| `event_prob()` | Probability of a named event by time *t* |
| `state_summary()` | Distribution of a state variable at each time point |
| `draws()` | Raw per-draw trajectories for detailed inspection |
| Parameter variation | Swap `delivery_bundle(params = ...)` to test scenarios |

**Next:** [03_decisions_policy.md](03_decisions_policy.md) — add decision points
and policies to the model, compare agent outcomes under different dispatch
strategies.
