---
title: "Cohort simulation and forecast"
maturity: "draft"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort simulation and forecast}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
suppressPackageStartupMessages({
  library(fluxCore)
  library(fluxForecast)
})
```

This tutorial shows how to go from a single-agent simulation (Tutorial 01) to a
full **fleet-level forecast**. By the end you will be able to:
- build a heterogeneous cohort of delivery agents,
- run the cohort through an 8-hour shift,
- generate probabilistic forecasts of delivery events and battery state across
  the fleet.

We use the urban food delivery model throughout. If you haven't worked through
[01_core_engine_scaffold.md](01_core_engine_scaffold.md), start there — it
covers the `Entity`, `ModelBundle`, and `Engine` concepts that this tutorial
builds on.

## Load the model

The delivery model lives in a plain R script. Sourcing it gives you
`delivery_bundle()`, `delivery_schema()`, and the individual callback functions.
No package install required.

```{r}
source("tutorials/model/urban_delivery.R")
```

## Build a heterogeneous cohort

A **cohort** is just a named list of `Entity` objects. Each agent starts its
shift with a different battery level, home zone, and starting state — exactly
the kind of heterogeneity you see in a real fleet.

```{r}
set.seed(2026)

n_agents <- 20
shared_schema <- delivery_schema()

agents <- lapply(seq_len(n_agents), function(i) {
  Entity$new(
    id   = paste0("driver_", sprintf("%02d", i)),
    init = list(
      battery_pct   = runif(1, min = 50, max = 100),
      route_zone    = sample(c("urban", "suburban", "rural"), 1,
                             prob = c(0.55, 0.30, 0.15)),
      payload_kg    = 0,
      dispatch_mode = "idle"
    ),
    schema      = shared_schema,
    entity_type = "delivery_agent",
    time0       = 0
  )
})
names(agents) <- vapply(agents, function(e) e$id, character(1))
```

Quick sanity check — the fleet's starting battery distribution:

```{r}
batteries <- vapply(agents, function(e) e$current$battery_pct, numeric(1))
summary(batteries)
```

## Single-agent run

Before running the full cohort, let's step through one agent to see the shape
of the output. The `Engine` is constructed once from the bundle and then reused
for every agent.

```{r}
eng <- Engine$new(bundle = delivery_bundle())

out_single <- eng$run(agents[[1]], max_events = 500, return_observations = TRUE)
```

The result contains:
- `$events` — the authoritative event log (one row per realized event)
- `$observations` — whatever the `observe()` hook emitted
- `$entity` — the same entity object, now mutated to post-run state

```{r}
nrow(out_single$events)
tail(out_single$observations, 5)
out_single$entity$state(c("battery_pct", "dispatch_mode"))
```

The final event should be `end_shift` — the model's terminal event. The battery
will be lower than it started, and the agent may have completed several
deliveries during the shift.

## Cohort simulation

`run_cohort()` runs the engine over every entity in the list, optionally with
multiple parameter draws (for uncertainty quantification) and multiple simulation
replicates per draw.

```{r}
cohort_result <- run_cohort(
  eng,
  entities     = agents,
  n_param_draws = 1,
  n_sims       = 50,
  max_events   = 500,
  seed         = 42
)
```

The result is an indexed list of run outputs. `cohort_result$index` tells you
which entity/draw/sim each slot corresponds to.

```{r}
head(cohort_result$index)
```

## Forecasting

Now the real value of the ecosystem: `fluxForecast` takes the raw simulation
output and answers probabilistic questions about the future.

### `forecast()` — the entry point

`forecast()` wraps the cohort run into a forecast object that downstream
functions can query. It needs the engine, the entities, and the evaluation
times (the "horizon grid" at which we want predictions).

```{r}
times <- seq(1, 8, by = 1)  # evaluate at hours 1 through 8

fc <- forecast(
  engine   = eng,
  entities = agents,
  times    = times,
  S        = 50,       # simulation draws per entity
  seed     = 42
)
```

### `event_prob()` — probability of a delivery event

"What is the probability that each agent completes at least one delivery by
hour *t*?"

```{r}
ep <- event_prob(fc, event = "delivery_completed", times = times)
head(ep)
```

Each row gives an entity × time combination with the estimated probability. You
can think of this as an "event risk" curve — it starts near 0 at t=0 and
approaches 1 as the shift progresses (most agents do complete deliveries).

Agents starting with low battery tend to have lower delivery probability early
on:

```{r}
# Compare the 5 lowest-battery agents vs the 5 highest
low_bat  <- names(sort(batteries))[1:5]
high_bat <- names(sort(batteries, decreasing = TRUE))[1:5]

ep_low  <- ep[ep$entity_id %in% low_bat & ep$time == 4, ]
ep_high <- ep[ep$entity_id %in% high_bat & ep$time == 4, ]

cat("Mean P(delivery by hour 4) — low battery group: ",
    round(mean(ep_low$prob), 3), "\n")
cat("Mean P(delivery by hour 4) — high battery group:",
    round(mean(ep_high$prob), 3), "\n")
```

### `state_summary()` — distribution of a state variable

"What does the battery distribution look like across the fleet at each hour?"

```{r}
ss <- state_summary(fc, vars = "battery_pct", times = times)
head(ss)
```

This gives you quantiles (or other summary statistics) of `battery_pct` at each
evaluation time. As the shift progresses, the distribution shifts left — agents
drain their batteries at different rates depending on their dispatch/delivery
intensity.

### `draws()` — inspect raw trajectories

For a single agent, you can pull the underlying simulation draws to see the
stochastic spread:

```{r}
dr <- draws(fc, var = "battery_pct", times = times)
# Filter to one agent
dr_one <- dr[dr$entity_id == "driver_01", ]
head(dr_one, 10)
```

Each row is one draw × one time point. You get the full distribution rather than
just a summary — useful for checking whether the forecast is well-behaved or if
there are pathological outliers.

## Varying model parameters

One of the design goals of the bundle architecture is that you can swap
parameters without changing any other code. Let's compare the default dispatch
rate against a slower fleet:

```{r}
# Default: dispatch_rate_base = 0.7
eng_slow <- Engine$new(bundle = delivery_bundle(
  params = list(dispatch_rate_base = 0.3)
))

fc_slow <- forecast(
  engine   = eng_slow,
  entities = agents,
  times    = times,
  S        = 50,
  seed     = 42
)

ep_slow <- event_prob(fc_slow, event = "delivery_completed", times = times)
```

Compare mean fleet-wide delivery probability at hour 4:

```{r}
cat("Default dispatch rate — P(delivery by hour 4):",
    round(mean(ep[ep$time == 4, "prob"]), 3), "\n")
cat("Slow dispatch rate   — P(delivery by hour 4):",
    round(mean(ep_slow[ep_slow$time == 4, "prob"]), 3), "\n")
```

The slower dispatch rate produces a flatter event probability curve — agents
receive fewer assignments, so fewer deliveries are completed by any given hour.
This is exactly the kind of "what if" scenario that fleet operators care about:
if demand drops (lower dispatch rate), how does delivery throughput change?

## Summary

| Concept | What you learned |
|---------|-----------------|
| `Entity` cohort | A named list of entities with heterogeneous starting state |
| `run_cohort()` | Batch simulation with parameter draws and replicates |
| `forecast()` | Wraps cohort output into a queryable forecast object |
| `event_prob()` | Probability of a named event by time *t* |
| `state_summary()` | Distribution of a state variable at each time point |
| `draws()` | Raw per-draw trajectories for detailed inspection |
| Parameter variation | Swap `delivery_bundle(params = ...)` to test scenarios |

**Next:** [04_decisions_policy.md](04_decisions_policy.md) — add decision points
and policies to the model, compare agent outcomes under different dispatch
strategies.
