---
title: "PROVISIONAL — Decision points and the action lifecycle"
maturity: "provisional"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PROVISIONAL - Decision points and the action lifecycle}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

> **Provisional, non-canonical, and intentionally unlinked.** This draft explores
> a possible future expansion of Tutorial 03. It is review material, not part of
> the supported tutorial sequence.

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
suppressPackageStartupMessages({
  library(fluxCore)
  library(knitr)
})
```

# One complete decision-to-action cycle

A decision point is a checkpoint attached to an event. When that event occurs,
the engine updates the entity, checks the decision point's condition, and asks
the policy for one `ActionEvent` or `NULL`. A proposed action joins the same
timeline as model events and is realized when its time comes.

This minimal deterministic model starts with one dispatch offer. The
`dispatch_check` transition records that an offer was seen. The policy then
schedules `accept`, whose action handler records the decision's effect.

```{r}
cycle_vars <- list(
  offered  = list(type = "count", default = 0L),
  accepted = list(type = "count", default = 0L)
)

cycle_dp <- DecisionPoint(
  id = "dispatch_decision",
  trigger = "dispatch_check",
  allowed_actions = "accept",
  action_handlers = list(
    accept = function(entity, event) list(accepted = 1L)
  )
)

cycle_schema <- set_schema(
  vars = cycle_vars,
  time_spec = time_spec(unit = "hours"),
  decision_points = list(cycle_dp)
)

cycle_bundle <- list(
  time_spec = time_spec(unit = "hours"),
  event_catalog = "dispatch_check",
  propose_events = function(entity) {
    if (as.integer(entity$current$offered) > 0L) return(list())
    list(dispatch = list(time_next = 1, event_type = "dispatch_check"))
  },
  transition = function(entity, event) {
    if (identical(event$event_type, "dispatch_check")) {
      return(list(offered = 1L))
    }
    list()
  },
  stop = function(entity, event) identical(event$event_type, "accept")
)

accept_policy <- function(decision_point, entity) {
  ActionEvent("accept", time_next = entity$last_time + 0.25)
}

cycle_engine <- load_model(
  schema = cycle_schema,
  bundle = cycle_bundle,
  policy = accept_policy,
  trajectory = list(detail = "full")
)

cycle_out <- cycle_engine$run(
  Entity$new(init = list(offered = 0L, accepted = 0L), schema = cycle_vars),
  max_events = 10
)

cycle_out$events |> kable()
cat("Accepted:", cycle_out$entity$current$accepted, "\n")
cat("Stopped by:", cycle_out$stopped_by, "\n")
```

The model has no process proposal after `dispatch_check`, but the pending
`accept` action still fires. Model proposals and policy actions are held
separately: refreshing or emptying the model's proposals cannot erase a pending
action.

# The lifecycle contract

- A policy call returns exactly one `ActionEvent` or `NULL`, never a list.
- Each decision point owns one pending-action slot. Different decision points
  can each have an action pending at the same time.
- Model refresh rules rewrite model proposals only. Pending actions survive.
- The engine retires an action when it is realized; model code never names or
  clears it.
- A realized action is an ordinary event. It can update state, be observed, and
  trigger another decision point.
- `action_handler()` replaces `transition()` for that action event. It is not an
  additional transition.

# Repeated proposals from one decision point

If a decision point fires again while its earlier action is still pending, its
`on_pending_action` setting controls the conflict:

| Mode | Result |
|---|---|
| `"warn"` | Replace the pending action and warn; this is the default |
| `"replace"` | Replace it silently, explicitly declaring that superseding is intended |
| `"keep"` | Keep the pending action and discard the new proposal |
| `"error"` | Stop with an error |

The following model receives a tick every hour but schedules its response three
hours out, guaranteeing a conflict on the second tick.

```{r}
pending_demo <- function(mode = NULL) {
  dp_args <- list(
    id = "reminder_decision",
    trigger = "tick",
    allowed_actions = "reminder",
    action_handlers = list(reminder = function(entity, event) list())
  )
  if (!is.null(mode)) dp_args$on_pending_action <- mode
  dp <- do.call(DecisionPoint, dp_args)

  vars <- list(ticks = list(type = "count", default = 0L))
  schema <- set_schema(
    vars = vars,
    time_spec = time_spec(unit = "hours"),
    decision_points = list(dp)
  )
  bundle <- list(
    time_spec = time_spec(unit = "hours"),
    event_catalog = "tick",
    propose_events = function(entity) {
      list(clock = list(time_next = entity$last_time + 1, event_type = "tick"))
    },
    transition = function(entity, event) {
      list(ticks = as.integer(entity$current$ticks) + 1L)
    },
    stop = function(entity, event) FALSE
  )
  policy <- function(decision_point, entity) {
    ActionEvent("reminder", time_next = entity$last_time + 3)
  }

  warnings <- character()
  value <- tryCatch(
    withCallingHandlers(
      load_model(schema, bundle, policy)$run(
        Entity$new(init = list(ticks = 0L), schema = vars),
        max_events = 2
      ),
      warning = function(w) {
        warnings <<- c(warnings, conditionMessage(w))
        invokeRestart("muffleWarning")
      }
    ),
    error = function(e) e
  )

  data.frame(
    mode = if (is.null(mode)) "warn (default)" else mode,
    outcome = if (inherits(value, "error")) "error" else value$stopped_by,
    warned = length(warnings) > 0L
  )
}

do.call(rbind, lapply(list(NULL, "replace", "keep", "error"), pending_demo)) |>
  kable()
```

Re-proposing immediately after the earlier action fired is not a conflict: the
engine retires the realized action before merging the new proposal.

# Alternatives versus sequences

Several decision points may trigger on the same event. Their actions occupy
independent slots, so scheduling one earlier does not cancel another—it only
determines which happens first.

When decision points describe alternative responses, make their conditions
mutually exclusive:

```{r, eval=FALSE}
routine <- DecisionPoint(
  id = "routine",
  trigger = "dispatch_check",
  condition = function(entity) entity$current$battery_pct >= 10
)
critical <- DecisionPoint(
  id = "critical",
  trigger = "dispatch_check",
  condition = function(entity) entity$current$battery_pct < 10
)
```

When both actions are intended, distinct times create a sequence. This small
model proves that the later action remains pending after the earlier one fires.

```{r}
sequence_vars <- list(started = list(type = "count", default = 0L))

sequence_dps <- list(
  DecisionPoint(
    id = "early",
    trigger = "tick",
    allowed_actions = "early_action",
    action_handlers = list(early_action = function(entity, event) list())
  ),
  DecisionPoint(
    id = "late",
    trigger = "tick",
    allowed_actions = "late_action",
    action_handlers = list(late_action = function(entity, event) list())
  )
)

sequence_schema <- set_schema(
  vars = sequence_vars,
  time_spec = time_spec(unit = "hours"),
  decision_points = sequence_dps
)

sequence_bundle <- list(
  time_spec = time_spec(unit = "hours"),
  event_catalog = "tick",
  propose_events = function(entity) {
    if (as.integer(entity$current$started) > 0L) return(list())
    list(clock = list(time_next = 1, event_type = "tick"))
  },
  transition = function(entity, event) list(started = 1L),
  stop = function(entity, event) FALSE
)

sequence_policy <- function(decision_point, entity) {
  if (identical(decision_point$id, "early")) {
    return(ActionEvent("early_action", entity$last_time + 0.1))
  }
  ActionEvent("late_action", entity$last_time + 0.2)
}

sequence_out <- load_model(
  sequence_schema, sequence_bundle, sequence_policy
)$run(Entity$new(init = list(started = 0L), schema = sequence_vars), max_events = 5)

sequence_out$events |> kable()
```

# Chaining decision points

Because an action is an event, it may trigger the next decision point. Chaining
is the supported way to model a multi-step workflow while each policy call
continues to return at most one action.

Here, `start` triggers stage A, whose action triggers stage B, whose action
triggers stage C. The delays are deliberately meaningful rather than tiny
ordering offsets.

```{r}
chain_vars <- list(
  started = list(type = "count", default = 0L),
  completed = list(type = "count", default = 0L)
)

chain_dps <- list(
  DecisionPoint(id = "stage_a", trigger = "start", allowed_actions = "stage_b"),
  DecisionPoint(id = "stage_b", trigger = "stage_b", allowed_actions = "stage_c"),
  DecisionPoint(id = "stage_c", trigger = "stage_c", allowed_actions = "complete")
)

chain_schema <- set_schema(
  vars = chain_vars,
  time_spec = time_spec(unit = "days"),
  decision_points = chain_dps
)

chain_bundle <- list(
  time_spec = time_spec(unit = "days"),
  # These actions have no action handlers, so their event types must be listed.
  event_catalog = c("start", "stage_b", "stage_c", "complete"),
  propose_events = function(entity) {
    if (as.integer(entity$current$started) > 0L) return(list())
    list(starter = list(time_next = 1, event_type = "start"))
  },
  transition = function(entity, event) {
    if (identical(event$event_type, "start")) return(list(started = 1L))
    if (identical(event$event_type, "complete")) return(list(completed = 1L))
    list()
  },
  stop = function(entity, event) identical(event$event_type, "complete")
)

chain_policy <- function(decision_point, entity) {
  switch(
    decision_point$id,
    stage_a = ActionEvent("stage_b", entity$last_time + 1),
    stage_b = ActionEvent("stage_c", entity$last_time + 2),
    stage_c = ActionEvent("complete", entity$last_time + 3)
  )
}

chain_out <- load_model(chain_schema, chain_bundle, chain_policy)$run(
  Entity$new(init = list(started = 0L, completed = 0L), schema = chain_vars),
  max_events = 10
)

chain_out$events |> kable()
cat("Completed:", chain_out$entity$current$completed, "\n")
```

When an action type has no `action_handler`, declare it in
`bundle$event_catalog`; otherwise validation fails when the action reaches the
timeline. A cycle such as A → B → A can run forever. Use `max_events` or
`max_time`, then inspect `stopped_by` to detect that backstop.

# Passing action parameters into proposal logic

`propose_events()` may declare `last_event`. It receives `NULL` initially and
the complete realized event after each step, including an action's `params`,
`metadata`, and `decision_point_id`.

This lets an action reschedule a future process without adding a variable whose
only purpose is carrying a delay through entity state.

```{r}
parameter_vars <- list(
  requested = list(type = "count", default = 0L),
  followed_up = list(type = "count", default = 0L)
)

parameter_dp <- DecisionPoint(
  id = "followup_decision",
  trigger = "request",
  allowed_actions = "schedule_followup"
)

parameter_schema <- set_schema(
  vars = parameter_vars,
  time_spec = time_spec(unit = "hours"),
  decision_points = list(parameter_dp)
)

parameter_bundle <- list(
  time_spec = time_spec(unit = "hours"),
  event_catalog = c("request", "schedule_followup", "followup"),
  propose_events = function(entity, last_event = NULL) {
    if (as.integer(entity$current$requested) == 0L) {
      return(list(request = list(time_next = 1, event_type = "request")))
    }
    if (identical(last_event$event_type, "schedule_followup")) {
      return(list(followup = list(
        time_next = entity$last_time + last_event$params$delay_hours,
        event_type = "followup"
      )))
    }
    list()
  },
  transition = function(entity, event) {
    if (identical(event$event_type, "request")) return(list(requested = 1L))
    if (identical(event$event_type, "followup")) return(list(followed_up = 1L))
    list()
  },
  stop = function(entity, event) identical(event$event_type, "followup")
)

parameter_policy <- function(decision_point, entity) {
  ActionEvent(
    "schedule_followup",
    time_next = entity$last_time + 0.1,
    params = list(delay_hours = 24)
  )
}

parameter_out <- load_model(
  parameter_schema, parameter_bundle, parameter_policy
)$run(
  Entity$new(
    init = list(requested = 0L, followed_up = 0L),
    schema = parameter_vars
  ),
  max_events = 10
)

parameter_out$events |> kable()
cat("State variables:", paste(names(parameter_out$entity$current), collapse = ", "), "\n")
```

The follow-up occurs 24 hours after the scheduling action. There is no delay
field in the schema: the value travels with the action and reaches proposal
logic through `last_event$params`.

# Provisional takeaways

- Decision points hold slots, not queues.
- The earliest pending item happens next; later actions remain pending.
- Conditions express alternatives, while distinct event times express a
  sequence.
- Chained decision points express multi-step workflows without multi-action
  policy returns.
- Action parameters can influence future process timing without transport
  state.
- `stopped_by` makes simulation backstops visible in the result.

> End of provisional draft. This document is intentionally absent from the
> supported tutorial navigation.
