---
title: "Decision points and policy"
maturity: "draft"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Decision points and policy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
suppressPackageStartupMessages(library(fluxCore))
```

In [Tutorial 02](02_cohort_forecast.md), every courier followed the same
mechanistic process: dispatches arrived, deliveries completed, batteries
drained — all without anyone making choices. Real systems involve **decisions**: a dispatcher choosing whether to
accept or decline an assignment, a routing algorithm picking surge vs normal
mode, a fleet manager pulling a low-battery vehicle off the road.

This tutorial introduces **decision points** and **policies** — the mechanism
by which choices are injected into the simulation timeline. A **policy** is a
set of decision rules: given a courier's current state at a specific moment in
the simulation, it proposes an action such as “accept this dispatch” or
“decline”. By the end you will be able to:

- declare a decision point on the delivery schema,
- write a policy function that proposes actions,
- compare outcomes under two different policies with identical seeds,
- inspect trajectory records to see exactly why each decision was made.

## Setup

```{r}
library(fluxCore)
source("tutorials/model/urban_delivery.R")
set.seed(2026)
```

## The decision: accept or decline a dispatch

When a `dispatch_check` event fires, the courier currently accepts every
assignment unconditionally. But what if a fleet management system could
**decline** an assignment — for example, when battery is too low to safely
complete the delivery?

This is a natural decision point. Let's formalize it.

## Declaring a decision point

A **decision point** is a named checkpoint in the event timeline where the
simulation pauses to ask: *should something intervene here?* When a decision
point fires, the engine calls your policy function and gives it a chance to
propose an action — for example, "decline this dispatch" or "switch to surge
mode". If no policy is attached, the simulation proceeds as if the checkpoint
were not there.

Decision points are declared on the schema, not buried in transition logic.
This keeps the interface explicit and auditable.

```{r}
dp_dispatch <- DecisionPoint(
  id              = "dispatch_decision",
  trigger         = "dispatch_check",
  allowed_actions = c("accept", "decline"),
  label           = "Accept or decline an incoming dispatch assignment"
)
```

The fields:
- **`trigger`**: which event(s) cause this decision point to fire. The
  simplest form is a character string (or vector) matching one or more event
  type names. You can also supply a function `function(event)` that inspects
  the event and returns `TRUE` or `FALSE` — useful when you want to fire only
  for a subset of events of the same type (e.g., only expedited deliveries).
  The function receives the event as it was *before* the transition runs, so
  it sees fields like `event_type`, `time_next`, and any `metadata` attached
  by the event generator. To check courier state *after* the transition,
  use `condition` instead.

  ```r
  # Fire only when the event was flagged as expedited
  trigger = function(event) {
    event$event_type == "delivery_completed" &&
      isTRUE(event$metadata$expedited)
  }
  ```

- **`condition`**: an optional function `function(entity)` that is checked
  *after* the transition runs and the courier's state has been updated. If it
  returns `FALSE`, the policy is not called for that event — the simulation
  continues without intervention. Use this to restrict the decision point to
  situations where the courier's current state warrants action:

  ```r
  # Only consult the policy when battery is below 25 % after the dispatch
  condition = function(entity) entity$current$battery_pct < 25
  ```

  `trigger` and `condition` serve different purposes: `trigger` selects
  *which events* activate the checkpoint; `condition` checks *whether the
  current state* warrants calling the policy.

- **`audit`**: logical (default `FALSE`). When `TRUE`, the simulation logs a
  `TrajectoryRecord` even for cycles where `condition` suppressed the policy
  call. The record is flagged with `condition_met = FALSE` and
  `selected_action = NULL`, giving you a complete history of every checkpoint
  visit regardless of whether the policy ran.

- **`allowed_actions`**: what the policy is permitted to propose. Actions
  outside this set are rejected.

- **`label`**: human-readable documentation.

Now assemble a full schema object (variables + time_spec + decision points):

```{r}
schema_with_dp <- set_schema(
  vars            = delivery_schema(),
  time_spec       = time_spec(unit = "hours"),
  decision_points = list(dp_dispatch)
)
```

## Writing policies

A **policy** is a plain R list with a `propose_action` function. That function
receives the decision point object and the current courier state, and returns
an `ActionEvent` proposing what should happen next — or `NULL` for "no
intervention."

### Policy A: always accept

The simplest possible policy — accept every dispatch regardless of state.
This is equivalent to having no policy at all, and serves as a baseline.

```{r}
policy_always_accept <- list(

  propose_action = function(decision_point, entity) {
    ActionEvent(
      action_type       = "accept",
      time_next         = entity$last_time + 0.01,
      decision_point_id = decision_point$id
    )
  }
)
```

### Policy B: battery threshold

Decline if battery has dropped below 25%. The idea: a courier running low on
battery should stop taking new assignments and coast to shift end rather than
risk stranding mid-delivery.

```{r}
policy_battery_threshold <- list(
  propose_action = function(decision_point, entity) {
    battery <- as.numeric(entity$current$battery_pct)

    action <- if (!is.null(battery) && battery < 25) "decline" else "accept"

    ActionEvent(
      action_type       = action,
      time_next         = entity$last_time + 0.01,
      decision_point_id = decision_point$id,
      metadata          = list(battery_at_decision = battery)
    )
  }
)
```

## Extending the bundle to handle the "decline" action

When the policy proposes "decline", that action enters the event timeline as
an `ActionEvent` and flows through `transition()`. We need the transition
function to handle it — in this case, by reverting the dispatch (resetting
mode back to idle and dropping the payload that was just assigned).

```{r}
# Extended transition that also handles the "decline" action
delivery_transition_with_policy <- function(entity, event, param_ctx = NULL) {
  # Handle the decline action: undo the dispatch assignment

  if (identical(event$event_type, "decline")) {
    return(list(
      dispatch_mode = "idle",
      payload_kg    = 0
    ))
  }

  # "accept" action: no additional state change needed (dispatch already applied)
  if (identical(event$event_type, "accept")) {
    return(NULL)
  }

  # All other events: delegate to the standard transition
  delivery_transition(entity, event, param_ctx)
}
```

Now build a bundle that uses this extended transition:

```{r}
delivery_bundle_with_policy <- function(params = list()) {
  base <- delivery_bundle(params)
  base$transition <- delivery_transition_with_policy
  base$event_catalog <- c(base$event_catalog, "accept", "decline")
  base
}
```

## Assembling with `load_model()`

When decision points are present, `load_model()` is the recommended assembly
path. It validates that the schema, bundle, policy, and trajectory configuration
are mutually consistent before any run begins.

```{r}
model_accept <- load_model(
  schema     = schema_with_dp,
  bundle     = delivery_bundle_with_policy(),
  policy     = policy_always_accept,
  trajectory = list(detail = "summary")
)

model_threshold <- load_model(
  schema     = schema_with_dp,
  bundle     = delivery_bundle_with_policy(),
  policy     = policy_battery_threshold,
  trajectory = list(detail = "summary")
)
```

## Running both policies from the same seed

The power of the decision-point architecture: same courier, same seed, same
stochastic draws — but different policies yield different outcomes.

```{r}
courier <- Entity$new(
  id          = "courier_A",
  init        = list(
    battery_pct   = 80,
    route_zone    = "urban",
    payload_kg    = 0,
    dispatch_mode = "idle"
  ),
  schema      = delivery_schema(),
  entity_type = "courier",
  time0       = 0
)

# Run under always-accept
out_accept <- model_accept$run(courier, max_events = 500,
                               return_observations = TRUE, seed = 99)

# Fresh courier (same starting state)
courier2 <- Entity$new(
  id          = "courier_A",
  init        = list(
    battery_pct   = 80,
    route_zone    = "urban",
    payload_kg    = 0,
    dispatch_mode = "idle"
  ),
  schema      = delivery_schema(),
  entity_type = "courier",
  time0       = 0
)

# Run under battery-threshold
out_threshold <- model_threshold$run(courier2, max_events = 500,
                                     return_observations = TRUE, seed = 99)
```

## Comparing outcomes

```{r}
# Count deliveries completed under each policy
count_deliveries <- function(out) {
  sum(out$events$event_type == "delivery_completed", na.rm = TRUE)
}

cat("Always-accept policy:\n")
cat("  Deliveries completed:", count_deliveries(out_accept), "\n")
cat("  Final battery:       ", round(out_accept$entity$current$battery_pct, 1), "%\n\n")

cat("Battery-threshold policy:\n")
cat("  Deliveries completed:", count_deliveries(out_threshold), "\n")
cat("  Final battery:       ", round(out_threshold$entity$current$battery_pct, 1), "%\n")
```

The always-accept courier takes every dispatch and drains the battery more
aggressively — potentially completing more deliveries but at higher risk of
hitting critical battery levels. The threshold courier conserves energy by
declining late-shift dispatches when battery is low.

## Inspecting trajectory records

Every time a decision point fires, the engine emits a `TrajectoryRecord`
capturing what was observed, what was proposed, and what happened next. This
is the audit trail that enables policy comparison and counterfactual analysis.

```{r}
# How many decisions were made?
cat("Decisions (accept policy):   ", length(out_accept$trajectory_records), "\n")
cat("Decisions (threshold policy):", length(out_threshold$trajectory_records), "\n")
```

Inspect the structure of a single record:

```{r}
tr <- out_threshold$trajectory_records[[1]]
cat("Time:            ", tr$t, "\n")
cat("Decision point:  ", tr$decision_point_id, "\n")
cat("Action proposed: ", tr$selected_action$action_type, "\n")
cat("Battery before:  ", tr$state_before$battery_pct, "\n")
cat("Battery after:   ", tr$state_after$battery_pct, "\n")
```

Build a summary table across all records:

```{r}
tr_df <- trajectory_table(out_threshold$trajectory_records,
                          vars = c("battery_pct", "dispatch_mode"))
head(tr_df, 10)
```

Look for the moment the policy starts declining — that's where the battery
crosses the threshold and the courier switches from accepting everything to
coasting to end of shift.

## Separating the guard from the action: `condition`

In `policy_battery_threshold` above, the state check lives inside
`propose_action`. This works, but it conflates two concerns: *when should
this decision be made?* and *what action to take given the current state?*

The `condition` parameter on `DecisionPoint()` pulls the state guard out of
the policy and onto the checkpoint itself. When `condition` returns `FALSE`,
the policy is never called and the simulation continues normally. The policy
only runs when the condition is already met, so it can always propose the same
action without repeating the check:

```{r}
dp_dispatch_cond <- DecisionPoint(
  id              = "dispatch_decision",
  trigger         = "dispatch_check",
  condition       = function(entity) entity$current$battery_pct < 25,
  allowed_actions = c("decline"),
  audit           = TRUE,
  label           = "Decline dispatch when battery is critically low; log all visits"
)

schema_cond <- set_schema(
  vars            = delivery_schema(),
  time_spec       = time_spec(unit = "hours"),
  decision_points = list(dp_dispatch_cond)
)

# Policy is now unconditional: condition has already done the filtering.
policy_decline_only <- list(
  propose_action = function(decision_point, entity) {
    ActionEvent(
      action_type       = "decline",
      time_next         = entity$last_time + 0.01,
      decision_point_id = decision_point$id
    )
  }
)

model_cond <- load_model(
  schema     = schema_cond,
  bundle     = delivery_bundle_with_policy(),
  policy     = policy_decline_only,
  trajectory = list(detail = "summary")
)
```

Also, `audit = TRUE` is introduced here. When `audit` is set on a decision
point that has a `condition`, the simulation logs a record for **every**
checkpoint visit — not just the ones where the condition was met and the policy
ran. Records where the condition was not met carry `condition_met = FALSE` and
`selected_action = NULL`. This gives you a complete history of when the
condition was and was not satisfied:

Run on the same courier with the same seed:

```{r}
courier_cond <- Entity$new(
  id          = "courier_A",
  init        = list(battery_pct=80, route_zone="urban",
                     payload_kg=0, dispatch_mode="idle"),
  schema      = delivery_schema(),
  entity_type = "courier",
  time0       = 0
)

out_cond <- model_cond$run(courier_cond, max_events = 500,
                           return_observations = TRUE, seed = 99)
```

Because `audit = TRUE`, the trajectory records include every
`dispatch_check` visit — not just the ones where the policy was called:

```{r}
cond_flags <- sapply(out_cond$trajectory_records, function(tr) tr$condition_met)

cat("Total DP visits (all dispatch_check events):",
    length(out_cond$trajectory_records), "\n")
cat("Condition met  (battery < 25, policy called):",
    sum(isTRUE(cond_flags) | is.na(cond_flags) == FALSE & cond_flags, na.rm = TRUE), "\n")
cat("Vetoed         (battery >= 25, logged only): ",
    sum(!cond_flags, na.rm = TRUE), "\n")
```

The `condition_met` field on each record tells you which records had the
policy consulted (`TRUE`) and which were logged-but-skipped (`FALSE`):

```{r}
tr_cond_df <- trajectory_table(out_cond$trajectory_records,
                               vars = c("battery_pct", "dispatch_mode"))
# condition_met column is present when a condition is declared on the DP
head(tr_cond_df[, intersect(names(tr_cond_df),
                             c("t", "battery_pct", "dispatch_mode", "condition_met",
                               "selected_action"))], 8)
```

The two approaches — state check in `propose_action` vs. `condition` on the
`DecisionPoint` — produce the same behavioral outcomes. Use `condition` when
the guard is always the same regardless of policy, or when you want the
audit trail to capture every DP visit.

## Multiple decision points per event cycle

A schema may carry more than one decision point on the same trigger. Each
fires **independently**: the engine evaluates trigger, condition, and calls
the policy for each active DP in sequence. Multiple `ActionEvent`s enter
the proposal queue and **arbitration** picks the earliest `time_next`.

A common pattern is a tiered response: one DP handles routine accept/decline
decisions and a second DP handles an emergency override when battery reaches
a critical threshold:

```{r}
dp_standard <- DecisionPoint(
  id              = "dispatch_accept",
  trigger         = "dispatch_check",
  allowed_actions = c("accept", "decline"),
  label           = "Routine accept/decline decision"
)

dp_critical <- DecisionPoint(
  id              = "battery_critical",
  trigger         = "dispatch_check",
  condition       = function(entity) entity$current$battery_pct < 10,
  allowed_actions = c("decline"),
  audit           = TRUE,
  label           = "Emergency override: battery critically low"
)

schema_two_dps <- set_schema(
  vars            = delivery_schema(),
  time_spec       = time_spec(unit = "hours"),
  decision_points = list(dp_standard, dp_critical)
)
```

When battery is above 10%, only `dp_standard` is active and the routine
policy runs. When battery drops below 10%, **both** DPs are active:
`dp_standard` proposes (accept/decline), `dp_critical` also proposes
"decline". Arbitration picks the proposal with the smallest `time_next` — so
you can force the critical override to win by setting a slightly smaller
offset:

```{r}
policy_two_dps <- list(
  propose_action = function(decision_point, entity) {
    battery <- as.numeric(entity$current$battery_pct)

    if (decision_point$id == "battery_critical") {
      # Emergency override fires first (time_next offset smaller).
      return(ActionEvent(
        action_type       = "decline",
        time_next         = entity$last_time + 0.001,
        decision_point_id = decision_point$id,
        metadata          = list(reason = "battery_critical")
      ))
    }

    # Routine policy: accept unless battery < 25.
    action <- if (!is.null(battery) && battery < 25) "decline" else "accept"
    ActionEvent(
      action_type       = action,
      time_next         = entity$last_time + 0.01,
      decision_point_id = decision_point$id
    )
  }
)
```

With two DPs both proposing actions on the same event step, the proposal
with `time_next = last_time + 0.001` is enqueued first. When it fires, the
transition handles "decline" and sets the stop-criterion; the second
proposal from the routine DP becomes unreachable.

> **Arbitration rule:** When multiple `ActionEvent`s are in the proposal
> queue, the engine selects the one with the smallest `time_next`. If
> `time_next` values are equal, ordering is undefined. Design your
> `time_next` offsets to encode priority.

## Cohort-level comparison

Let's scale this up. Run a 20-courier cohort under both policies and compare
aggregate outcomes.

```{r}
make_couriers <- function(n = 20, seed = 2026) {
  set.seed(seed)
  lapply(seq_len(n), function(i) {
    Entity$new(
      id   = paste0("courier_", sprintf("%02d", i)),
      init = list(
        battery_pct   = runif(1, 50, 100),
        route_zone    = sample(c("urban", "suburban", "rural"), 1,
                               prob = c(0.55, 0.30, 0.15)),
        payload_kg    = 0,
        dispatch_mode = "idle"
      ),
      schema      = delivery_schema(),
      entity_type = "courier",
      time0       = 0
    )
  })
}

cohort_accept    <- lapply(make_couriers(), function(e) {
  model_accept$run(e, max_events = 500, return_observations = TRUE, seed = 42)
})

cohort_threshold <- lapply(make_couriers(), function(e) {
  model_threshold$run(e, max_events = 500, return_observations = TRUE, seed = 42)
})

# Aggregate
del_accept    <- vapply(cohort_accept, count_deliveries, integer(1))
del_threshold <- vapply(cohort_threshold, count_deliveries, integer(1))
bat_accept    <- vapply(cohort_accept,
                        function(o) o$entity$current$battery_pct, numeric(1))
bat_threshold <- vapply(cohort_threshold,
                        function(o) o$entity$current$battery_pct, numeric(1))

cat("Fleet summary — always accept:\n")
cat("  Mean deliveries:", round(mean(del_accept), 1), "\n")
cat("  Mean final battery:", round(mean(bat_accept), 1), "%\n\n")

cat("Fleet summary — battery threshold:\n")
cat("  Mean deliveries:", round(mean(del_threshold), 1), "\n")
cat("  Mean final battery:", round(mean(bat_threshold), 1), "%\n")
```

The trade-off is visible at fleet scale: the threshold policy sacrifices some
delivery throughput in exchange for better battery preservation — a real
operational consideration when battery replacement or charging infrastructure
is constrained.

## What trajectory records enable

`TrajectoryRecord` is not just an audit log. It is the structured surface that
the rest of the flux ecosystem builds on:

- **Policy comparison**: run the same courier under two policies with the same
  seed; diff the trajectory_records to see where and why they diverge.
- **Counterfactual analysis**: fix seed + parameter draw; vary only the policy;
  compare outcomes.
- **RL training** (future: `fluxSim`): the `(observation, action, reward,
  next_state)` tuple for each record becomes a training transition.
- **Audit and explainability**: for any individual run, reconstruct the full
  decision history and ask "why did the simulation do that?"

## Summary

| Concept | What you learned |
|---------|-----------------|
| `DecisionPoint()` | Declares where in the event timeline a policy is consulted |
| `trigger` | Pre-transition gate: char event name(s) or `function(event)` on event-level fields |
| `condition` | Post-transition guard: `function(entity)` on updated state; vetoed cycles skip the policy call |
| `audit = TRUE` | Emit `TrajectoryRecord` even when `condition` vetoed; `condition_met = FALSE` flags those records |
| Policy function | `propose_action(dp, entity)` → `ActionEvent` or NULL; add `sim_ctx` / `param_ctx` to the signature only when needed |
| `ActionEvent()` | The proposed intervention — enters the timeline like any other event |
| `load_model()` | Validates schema + bundle + policy + trajectory config together |
| Trajectory records | Per-decision audit trail: `observation`, `action`, `state_before/after`, `condition_met` |
| Same seed, different policy | Isolates the causal effect of the policy on outcomes |
| Multiple DPs + arbitration | Multiple DPs can fire in one event cycle; earliest `time_next` wins |

**Next:** [05_prepare_operational_data.md](05_prepare_operational_data.md) —
generate synthetic operational logs and prepare them into train/test/validation
format with `fluxPrepare`.
