---
title: "Decision points and policy"
maturity: "draft"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Decision points and policy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
suppressPackageStartupMessages(library(fluxCore))
```

In [Tutorial 02](02_cohort_forecast.md), every courier followed the same
mechanistic process: dispatches arrived, deliveries completed, batteries
drained — all without anyone making choices. Real systems involve **decisions**: a dispatcher choosing whether to
accept or decline an assignment, a routing algorithm picking surge vs normal
mode, a fleet manager pulling a low-battery vehicle off the road.

This tutorial introduces **decision points** and **policies** — the mechanism
by which choices are injected into the simulation timeline. A **policy** is a
function that looks at the courier's current state and proposes an action: accept
the dispatch (take the payload, spend battery to deliver it) or decline (stay
idle, conserve battery for later). Accepting means more deliveries completed
but faster battery depletion; declining preserves range at the cost of missed
work. The question we will answer: *does a simple battery-conservation policy
improve the delivery-vs-battery trade-off compared to always accepting?*

By the end you will be able to:

- declare a decision point on the delivery schema,
- write a policy function that proposes actions,
- compare outcomes under two different policies with identical seeds,
- inspect trajectory records to see exactly why each decision was made,
- view the distribution of counterfactual outcomes across many runs.

## Setup

```{r}
library(fluxCore)
source("tutorials/model/urban_delivery.R")
set.seed(2026)
```

## The decision: accept or decline a dispatch

Recall the model from Tutorial 01. The simulation runs in discrete steps:
each step, competing event streams (dispatch arrivals, delivery completions,
shift end) each propose a next-event time; the engine picks the soonest one,
advances the clock, and calls `transition()` to update state variables.
Among those events is `dispatch_check` — the moment a new delivery assignment
arrives and the courier is offered work.

In the base model, `dispatch_check` always results in the courier accepting:
the transition assigns a route, adds payload, and deducts battery. But what if
a fleet management system could **decline** an assignment — for example, when
battery is too low to safely complete the delivery?

This is a natural decision point. Let's formalize it.

## Declaring a decision point

A **decision point** is a named checkpoint in the event timeline where the
simulation pauses to ask: *should something intervene here?* When a decision
point fires, the engine calls your policy function and gives it a chance to
propose an action — for example, "decline this dispatch" or "switch to surge
mode". If no policy is attached, the simulation proceeds as if the checkpoint
were not there.

Decision points are declared on the schema, not buried in transition logic.
This keeps the interface explicit and auditable.

```{r}
dp_dispatch <- DecisionPoint(
  id              = "dispatch_decision",
  trigger         = "dispatch_check",
  allowed_actions = c("accept", "decline"),
  label           = "Accept or decline an incoming dispatch assignment"
)
```

The fields:
- **`trigger`**: which event(s) cause this decision point to fire. Here we use
  the string `"dispatch_check"`, meaning the checkpoint activates every time
  a dispatch arrives. You can also supply a vector of event names, or a
  function `function(event)` that inspects event metadata and returns
  `TRUE`/`FALSE` for selective firing.

- **`condition`**: an optional function `function(entity)` that is evaluated
  *after* the transition runs. Why after? Because the fleet manager needs
  up-to-date state to make a decision — the courier's `battery_pct` may have
  changed during the transition. If `condition` returns `FALSE`, the policy
  is skipped for this event cycle. We don't use a condition yet (every
  `dispatch_check` should prompt a decision), but we will introduce it later.

- **`audit`**: logical (default `FALSE`). When `TRUE`, the simulation logs a
  `TrajectoryRecord` even for cycles where `condition` suppressed the policy
  call — useful for post-hoc analysis of "what *could* have happened."

- **`allowed_actions`**: the set of actions the policy is permitted to
  propose. Anything outside this set is rejected by the engine.

Now assemble a full schema object (variables + time_spec + decision points):

```{r}
schema_with_dp <- set_schema(
  vars            = delivery_schema(),
  time_spec       = time_spec(unit = "hours"),
  decision_points = list(dp_dispatch)
)
```

## Writing policies

A **policy** is a plain R function that receives the firing decision point and
the current courier state, and returns an `ActionEvent` proposing what should
happen next — or `NULL` for "no intervention."

### Policy A: always accept

The simplest possible policy — accept every dispatch regardless of state.
This is equivalent to having no policy at all, and serves as a baseline.

```{r}
policy_always_accept <- function(decision_point, entity) {
  ActionEvent(
    action_type = "accept",
    time_next   = entity$last_time + 0.01
  )
}
```

### Policy B: battery threshold

Decline if battery has dropped below 60%. The idea: a courier running low on
battery should stop taking new assignments and coast to shift end rather than
risk stranding mid-delivery.

```{r}
policy_battery_threshold <- function(decision_point, entity) {
  battery <- as.numeric(entity$current$battery_pct)

  action <- if (!is.null(battery) && battery < 60) "decline" else "accept"

  ActionEvent(
    action_type = action,
    time_next   = entity$last_time + 0.01,
    metadata    = list(battery_at_decision = battery)
  )
}
```

## Extending the bundle to handle the "decline" action

When the policy proposes "decline", that action enters the event timeline as
an `ActionEvent` and flows through `transition()`. We need the transition
function to handle it — in this case, by reverting the dispatch (resetting
mode back to idle and dropping the payload that was just assigned).

There is one additional modelling choice here. In the base model, battery
drain is charged at `dispatch_check` — meaning the cost is incurred
regardless of whether the courier accepts. For the policy to have real
consequences, the battery cost must be moved to the `accept` event. A courier
who declines neither takes on payload nor pays the battery cost:

```{r}
# Extended transition: battery cost on "accept" (not on dispatch_check),
# so the policy choice has real consequences for remaining range.
delivery_transition_with_policy <- function(entity, event, param_ctx = NULL) {
  params <- if (is.list(param_ctx) && is.list(param_ctx$params)) param_ctx$params else list()
  pnum <- function(name, default) {
    x <- params[[name]]
    if (is.null(x) || length(x) != 1L) return(default)
    x <- suppressWarnings(as.numeric(x))
    if (!is.finite(x)) default else x
  }

  if (identical(event$event_type, "dispatch_check")) {
    # Offer arrives: assign route and payload only.  Battery cost comes later.
    route_probs <- params$route_zone_probs
    if (is.null(route_probs) || !is.numeric(route_probs) || length(route_probs) != 3L) {
      route_probs <- c(0.55, 0.30, 0.15)
    }
    return(list(
      route_zone    = sample(c("urban", "suburban", "rural"), size = 1,
                             prob = route_probs / sum(route_probs)),
      dispatch_mode = "assigned",
      payload_kg    = rlnorm(1,
        meanlog = log(max(0.1, pnum("dispatch_payload_mean_kg", 3.0))),
        sdlog   = max(0.05, pnum("dispatch_payload_sdlog", 0.35)))
    ))
  }

  if (identical(event$event_type, "accept")) {
    # Courier takes the job: pay the battery cost now.
    battery_now <- as.numeric(entity$current$battery_pct)
    if (!is.finite(battery_now)) battery_now <- 100
    battery_drop <- rexp(1, rate = 1 / max(0.1, pnum("dispatch_battery_drop_mean", 2.5)))
    return(list(battery_pct = max(0, battery_now - battery_drop)))
  }

  if (identical(event$event_type, "decline")) {
    # Courier passes: undo the assignment, no battery cost.
    return(list(dispatch_mode = "idle", payload_kg = 0))
  }

  delivery_transition(entity, event, param_ctx)
}
```

Now build a bundle that uses this extended transition. We also suppress
`delivery_completed` proposals while a courier is idle with no payload — a
consistency guard that prevents spurious delivery events:

```{r}
delivery_bundle_with_policy <- function(params = list()) {
  base          <- delivery_bundle(params)
  base_propose  <- base$propose

  base$transition   <- delivery_transition_with_policy
  base$event_catalog <- c(base$event_catalog, "accept", "decline")

  # Suppress delivery proposals when the courier has no payload.
  base$propose <- function(entity, param_ctx = NULL,
                           process_ids = NULL, current_proposals = NULL) {
    props   <- base_propose(entity, param_ctx, process_ids, current_proposals)
    payload <- suppressWarnings(as.numeric(entity$current$payload_kg))
    if (is.finite(payload) && payload <= 0) props$delivery <- NULL
    props
  }

  base
}
```

## Assembling with `load_model()`

When decision points are present, `load_model()` is the recommended assembly
path. It validates that the schema, bundle, policy, and trajectory configuration
are mutually consistent before any run begins.

Reproducibility is controlled through a `RuntimeContext`. Passing a `seed`
there ensures that every stochastic draw during the run — dispatch timing,
battery drain, route assignment — starts from a known state. Using the same
seed value across both models is what makes the comparison meaningful: every
difference in outcome is caused by the policy, not by random chance.

```{r}
model_accept <- load_model(
  schema     = schema_with_dp,
  bundle     = delivery_bundle_with_policy(),
  policy     = policy_always_accept,
  trajectory = list(detail = "summary"),
  runtime    = RuntimeContext(seed = 500)
)

model_threshold <- load_model(
  schema     = schema_with_dp,
  bundle     = delivery_bundle_with_policy(),
  policy     = policy_battery_threshold,
  trajectory = list(detail = "summary"),
  runtime    = RuntimeContext(seed = 500)
)
```

## Running both policies from the same seed

The power of the decision-point architecture: same courier, same seed, same
stochastic draws — but different policies yield different outcomes. We create
two identical couriers (same starting state) and run each through its
respective model.

```{r}
courier <- Entity$new(
  id          = "courier_A",
  init        = list(
    battery_pct   = 80,
    route_zone    = "urban",
    payload_kg    = 0,
    dispatch_mode = "idle"
  ),
  schema      = delivery_schema(),
  entity_type = "courier",
  time0       = 0
)

# Fresh courier with identical starting state for the second model
courier2 <- Entity$new(
  id          = "courier_A",
  init        = list(
    battery_pct   = 80,
    route_zone    = "urban",
    payload_kg    = 0,
    dispatch_mode = "idle"
  ),
  schema      = delivery_schema(),
  entity_type = "courier",
  time0       = 0
)

out_accept    <- model_accept$run(courier,  max_events = 500, return_observations = TRUE)
out_threshold <- model_threshold$run(courier2, max_events = 500, return_observations = TRUE)
```

## Comparing outcomes

Let's look at the headline numbers: how many deliveries did each courier
complete, and how much battery was left at the end of the shift?

```{r}
# Count deliveries completed under each policy
count_deliveries <- function(out) {
  sum(out$events$event_type == "delivery_completed", na.rm = TRUE)
}

cat("Always-accept policy:\n")
cat("  Deliveries completed:", count_deliveries(out_accept), "\n")
cat("  Final battery:       ", round(out_accept$entity$current$battery_pct, 1), "%\n\n")

cat("Battery-threshold policy:\n")
cat("  Deliveries completed:", count_deliveries(out_threshold), "\n")
cat("  Final battery:       ", round(out_threshold$entity$current$battery_pct, 1), "%\n")
```

The always-accept courier takes every dispatch and drains the battery more
aggressively — potentially completing more deliveries but at higher risk of
hitting critical battery levels. The threshold courier conserves energy by
declining late-shift dispatches when battery is low.

## Inspecting trajectory records

Every time a decision point fires, the engine records what happened in a
`TrajectoryRecord`. Think of it as the simulation's decision log: it captures
the courier's state at the moment of the decision, what the policy proposed,
and what the state looked like afterward. This is what allows you to go back
after a run and ask *why* a particular decision was made.

```{r}
# How many decisions were made?
cat("Decisions (accept policy):   ", length(out_accept$trajectory_records), "\n")
cat("Decisions (threshold policy):", length(out_threshold$trajectory_records), "\n")
```

Inspect a single record to see the structure:

```{r}
tr <- out_threshold$trajectory_records[[1]]
cat("Time:            ", tr$t, "\n")
cat("Decision point:  ", tr$decision_point_id, "\n")
cat("Action proposed: ", tr$selected_action$action_type, "\n")
cat("Battery before:  ", tr$state_before$battery_pct, "\n")
cat("Battery after:   ", tr$state_after$battery_pct, "\n")
```

`state_before` and `state_after` bracket the *trigger event* (`dispatch_check`).
In this model battery is charged at the subsequent `accept` event, so both
readings show the same value unless battery changed for another reason during the
transition. The battery drop appears only in the record for the `accept` event
itself, which is not a decision point — it is purely mechanical.

`trajectory_table()` collects all records into a data frame, with one row per
decision and columns for time, state variables, and the action taken. Look for
the moment the policy starts declining — that's where the battery crosses 60%
and the courier switches from accepting everything to coasting to end of shift.

```{r}
tr_df <- trajectory_table(out_threshold$trajectory_records,
                          vars = c("battery_pct", "dispatch_mode"))
head(tr_df, 10)
```

## Using `condition` to restrict when the policy runs

In `policy_battery_threshold` above, the battery check lives inside the
policy function itself. That’s fine for simple cases, but it conflates two
concerns: *when* to intervene and *what* to do. If you always want to decline
when battery is critically low — regardless of which policy is plugged in —
you can move the guard out of the policy and onto the decision point.

The `condition` parameter accepts a function `function(entity)` that is
evaluated after the transition (so it sees current state). If it returns
`FALSE`, the policy is never called for that cycle.

Here we set a stricter threshold (25%) as a hard rule on the decision point.
The policy itself becomes trivial — it always proposes "decline" because it
only runs when the condition has already determined that intervention is
warranted:

```{r}
dp_dispatch_cond <- DecisionPoint(
  id              = "dispatch_decision",
  trigger         = "dispatch_check",
  condition       = function(entity) entity$current$battery_pct < 25,
  allowed_actions = c("decline"),
  audit           = TRUE,
  label           = "Decline dispatch when battery is critically low; log all visits"
)

schema_cond <- set_schema(
  vars            = delivery_schema(),
  time_spec       = time_spec(unit = "hours"),
  decision_points = list(dp_dispatch_cond)
)

# Policy is now unconditional: the condition handles the filtering,
# so every time this policy is called, the answer is always "decline".
policy_decline_only <- function(decision_point, entity) {
  ActionEvent(
    action_type = "decline",
    time_next   = entity$last_time + 0.01
  )
}

model_cond <- load_model(
  schema     = schema_cond,
  bundle     = delivery_bundle_with_policy(),
  policy     = policy_decline_only,
  trajectory = list(detail = "summary"),
  runtime    = RuntimeContext(seed = 99)
)
```

Also, `audit = TRUE` is introduced here. Normally, when `condition` returns
`FALSE` — meaning the battery is still healthy — the checkpoint fires and is
immediately skipped with no record kept. With `audit = TRUE`, a record is
written for *every* `dispatch_check` visit, whether or not the condition was
met. Records where the policy was suppressed carry `condition_met = FALSE`
and `selected_action = NULL`. This gives you a full picture of when couriers
were and were not at risk.

Run on the same courier with the same seed:

```{r}
courier_cond <- Entity$new(
  id          = "courier_A",
  init        = list(battery_pct=80, route_zone="urban",
                     payload_kg=0, dispatch_mode="idle"),
  schema      = delivery_schema(),
  entity_type = "courier",
  time0       = 0
)

out_cond <- model_cond$run(courier_cond, max_events = 500,
                           return_observations = TRUE)
```

Because `audit = TRUE`, the trajectory records include every
`dispatch_check` visit — not just the ones where the policy was called:

```{r}
cond_flags <- sapply(out_cond$trajectory_records, function(tr) tr$condition_met)

cat("Total DP visits (all dispatch_check events):",
    length(out_cond$trajectory_records), "\n")
cat("Condition met  (battery < 25, policy called):",
    sum(cond_flags == TRUE, na.rm = TRUE), "\n")
cat("Vetoed         (battery >= 25, logged only):  ",
    sum(cond_flags == FALSE, na.rm = TRUE), "\n")
```

The `condition_met` field in each record tells you which visits had the policy
consulted (`TRUE`) and which were logged but skipped (`FALSE`). The
`trajectory_table()` helper brings this into a data frame alongside the state
variables you care about:

```{r}
tr_cond_df <- trajectory_table(out_cond$trajectory_records,
                               vars = c("battery_pct", "dispatch_mode"))
head(tr_cond_df[, intersect(names(tr_cond_df),
                             c("t", "battery_pct", "dispatch_mode", "condition_met",
                               "selected_action"))], 8)
```

Rows with `condition_met = FALSE` are the visits where battery was still above
25% and no action was needed. Rows with `condition_met = TRUE` are the ones
where the policy ran and proposed "decline".

The two approaches — state check in the policy function vs. `condition` on the
`DecisionPoint` — produce identical behavioral outcomes. Use `condition` when
the guard is always the same regardless of policy, or when you want the
audit trail to capture every checkpoint visit.

## Multiple decision points per event cycle

A schema can carry more than one decision point attached to the same trigger.
Each fires **independently**: the engine checks the condition (if any) and
calls the policy separately for each active checkpoint in the same event step.
When both propose actions, both `ActionEvent`s enter a queue and
**arbitration** picks the one with the earliest `time_next`.

A common pattern is a tiered response: one checkpoint handles the routine
accept/decline decision, and a second handles an emergency override when the
battery reaches a critical threshold. Because the critical override should
always win, it is given a slightly smaller `time_next` offset so it is
scheduled ahead of the routine action:

```{r}
dp_standard <- DecisionPoint(
  id              = "dispatch_accept",
  trigger         = "dispatch_check",
  allowed_actions = c("accept", "decline"),
  label           = "Routine accept/decline decision"
)

dp_critical <- DecisionPoint(
  id              = "battery_critical",
  trigger         = "dispatch_check",
  condition       = function(entity) entity$current$battery_pct < 10,
  allowed_actions = c("decline"),
  audit           = TRUE,
  label           = "Emergency override: battery critically low"
)

schema_two_dps <- set_schema(
  vars            = delivery_schema(),
  time_spec       = time_spec(unit = "hours"),
  decision_points = list(dp_standard, dp_critical)
)
```

When battery is above 10%, only `dp_standard` is active and the routine
policy runs. When battery drops below 10%, **both** checkpoints fire:
`dp_standard` proposes its action and `dp_critical` also proposes "decline".
Arbitration picks whichever has the smaller `time_next`. By giving the
critical override a smaller offset (`+ 0.001` vs. `+ 0.01`), it always wins
when both are active:

```{r}
policy_two_dps <- function(decision_point, entity) {
  battery <- as.numeric(entity$current$battery_pct)

  if (decision_point$id == "battery_critical") {
    # Emergency override fires first (time_next offset smaller).
    return(ActionEvent(
      action_type = "decline",
      time_next   = entity$last_time + 0.001,
      metadata    = list(reason = "battery_critical")
    ))
  }

  # Routine policy: accept unless battery < 25.
  action <- if (!is.null(battery) && battery < 25) "decline" else "accept"
  ActionEvent(
    action_type = action,
    time_next   = entity$last_time + 0.01
  )
}
```

With two checkpoints both proposing actions in the same event step, the one
with `time_next = last_time + 0.001` is scheduled first. The transition
handles "decline" and the run continues; the second proposal from the routine
checkpoint has no further effect.

> **Arbitration rule:** When multiple proposed actions are waiting, the engine
> picks the one scheduled earliest (smallest `time_next`). If two proposals
> share the same time, the ordering is not defined — encode priority by
> choosing distinct offsets.

## Distribution of outcomes across many runs

One seed is one sample path. To see the real trade-off, run the same courier
500 times under each policy — each replicate gets its own seed, so the
stochastic draws (dispatch timing, battery drain, route) vary across runs, but
the comparison between policies is exact within each replicate because both
use the same seed.

```{r}
make_fresh_courier <- function() {
  Entity$new(
    id          = "courier_A",
    init        = list(battery_pct = 80, route_zone = "urban",
                       payload_kg = 0, dispatch_mode = "idle"),
    schema      = delivery_schema(),
    entity_type = "courier",
    time0       = 0
  )
}

n_reps <- 500

dist_results <- lapply(seq_len(n_reps), function(s) {
  m_a <- load_model(
    schema  = schema_with_dp, bundle = delivery_bundle_with_policy(),
    policy  = policy_always_accept, trajectory = list(detail = "none"),
    runtime = RuntimeContext(seed = s)
  )
  m_t <- load_model(
    schema  = schema_with_dp, bundle = delivery_bundle_with_policy(),
    policy  = policy_battery_threshold, trajectory = list(detail = "none"),
    runtime = RuntimeContext(seed = s)
  )
  oa <- m_a$run(make_fresh_courier(), max_events = 500)
  ot <- m_t$run(make_fresh_courier(), max_events = 500)
  list(
    del_accept    = count_deliveries(oa),
    del_threshold = count_deliveries(ot),
    bat_accept    = oa$entity$current$battery_pct,
    bat_threshold = ot$entity$current$battery_pct
  )
})
```

Because both policies run with the same seed in each replicate, we can compute
the *within-replicate* difference — the pure causal effect of switching policy,
with all randomness held constant:

```{r}
del_a   <- sapply(dist_results, `[[`, "del_accept")
del_t   <- sapply(dist_results, `[[`, "del_threshold")
bat_a   <- sapply(dist_results, `[[`, "bat_accept")
bat_t   <- sapply(dist_results, `[[`, "bat_threshold")

# Within-replicate differences
delta_del <- del_t - del_a   # > 0: threshold got more deliveries
delta_bat <- bat_t - bat_a   # > 0: threshold preserved more battery

cat("Deliveries — always-accept:    mean =", round(mean(del_a), 1),
    " median =", median(del_a), "\n")
cat("Deliveries — battery threshold: mean =", round(mean(del_t), 1),
    " median =", median(del_t), "\n")
cat("\n")
cat("Within-replicate delta (threshold minus accept):\n")
cat("  Deliveries: mean =", round(mean(delta_del), 2),
    "  P(threshold > accept) =",
    round(mean(delta_del > 0), 2), "\n")
cat("  Battery:    mean =", round(mean(delta_bat), 1), "%",
    "  P(threshold > accept) =",
    round(mean(delta_bat > 0), 2), "\n")
```

There is a genuine trade-off — and in a non-trivial fraction of runs, the
threshold policy actually delivers *more* than the always-accept policy. This
happens when greedily accepting every dispatch early in the shift burns battery
that would have been needed for dispatches later. The distribution of
counterfactual outcomes is the right object to reason about, not a single
point estimate.

```{r}
# Tabulate the distribution of delivery counts under each policy
tab_accept    <- table(factor(del_a,    levels = 0:max(c(del_a, del_t))))
tab_threshold <- table(factor(del_t, levels = 0:max(c(del_a, del_t))))

cat("Distribution of deliveries per shift (n =", n_reps, "replicates):\n\n")
cat(sprintf("%-12s  %8s  %8s\n", "Deliveries", "Accept", "Thresh"))
for (k in names(tab_accept)) {
  cat(sprintf("%-12s  %8d  %8d\n", k, tab_accept[[k]], tab_threshold[[k]]))
}
```

## What trajectory records enable

`TrajectoryRecord` is not just an audit log. It is the structured surface that
the rest of the flux ecosystem builds on:

- **Policy comparison**: run the same courier under two policies with the same
  seed; diff the trajectory_records to see where and why they diverge.
- **Counterfactual analysis**: fix seed + parameter draw; vary only the policy;
  compare outcomes.
- **RL training** (future: `fluxSim`): the `(observation, action, reward,
  next_state)` tuple for each record becomes a training transition.
- **Audit and explainability**: for any individual run, reconstruct the full
  decision history and ask "why did the simulation do that?"

## Summary

| Concept | What you learned |
|---------|-----------------|
| `DecisionPoint()` | Declares where in the event timeline a policy is consulted |
| `trigger` | Pre-transition gate: char event name(s) or `function(event)` on event-level fields |
| `condition` | Post-transition guard: `function(entity)` on updated state; vetoed cycles skip the policy call |
| `audit = TRUE` | Emit `TrajectoryRecord` even when `condition` vetoed; `condition_met = FALSE` flags those records |
| Policy function | `function(decision_point, entity)` → `ActionEvent` or NULL; add `sim_ctx` / `param_ctx` to the signature only when needed |
| `ActionEvent()` | The proposed intervention — enters the timeline like any other event |
| `load_model()` | Validates schema + bundle + policy + trajectory config together |
| Trajectory records | Per-decision audit trail: `observation`, `action`, `state_before/after`, `condition_met` |
| Same seed, different policy | Isolates the causal effect of the policy on outcomes |
| Multiple DPs + arbitration | Multiple DPs can fire in one event cycle; earliest `time_next` wins |

**Next:** [05_prepare_operational_data.md](05_prepare_operational_data.md) —
generate synthetic operational logs and prepare them into train/test/validation
format with `fluxPrepare`.
