---
title: "Decision points and policy"
maturity: "draft"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Decision points and policy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
suppressPackageStartupMessages(library(fluxCore))
```

In Tutorial 03, every agent followed the same mechanistic process: dispatches
arrived, deliveries completed, batteries drained — all without anyone making
choices. Real systems involve **decisions**: a dispatcher choosing whether to
accept or decline an assignment, a routing algorithm picking surge vs normal
mode, a fleet manager pulling a low-battery vehicle off the road.

This tutorial introduces **decision points** — the mechanism by which policies
inject choices into the simulation timeline. By the end you will be able to:

- declare a decision point on the delivery schema,
- write a policy function that proposes actions,
- compare outcomes under two different policies with identical seeds,
- inspect trajectory records to see exactly why each decision was made.

## Setup

```{r}
library(fluxCore)
source("tutorials/model/urban_delivery.R")
set.seed(2026)
```

## The decision: accept or decline a dispatch

When a `dispatch_check` event fires, the agent currently accepts every
assignment unconditionally. But what if the agent (or a fleet management system)
could **decline** an assignment — for example, when battery is too low to
safely complete the delivery?

This is a natural decision point. Let's formalize it.

## Declaring a decision point

Decision points are declared on the schema, not buried in transition logic.
This makes the policy interface explicit and auditable. A schema without
`$decision_points` runs as a pure generative model — policies are opt-in.

```{r}
dp_dispatch <- DecisionPoint(
  id              = "dispatch_decision",
  trigger         = "dispatch_check",
  allowed_actions = c("accept", "decline"),
  label           = "Accept or decline an incoming dispatch assignment"
)
```

The fields:
- **`trigger`**: the event type that fires this decision point. The engine
  invokes the policy *after* transition has already been applied for that event.
- **`allowed_actions`**: what the policy is permitted to propose. Actions
  outside this set are rejected.
- **`label`**: human-readable documentation.

Now assemble a full schema object (variables + time_spec + decision points):

```{r}
schema_with_dp <- list(
  variables       = delivery_schema(),
  time_spec       = time_spec(unit = "hours"),
  decision_points = list(dp_dispatch)
)
```

## Writing policies

A policy is a list with a `propose_action` method:

```r
propose_action(decision_point, entity, sim_ctx, param_ctx) -> ActionEvent | NULL
```

It inspects the current entity state (which already reflects the triggering
event's transition) and returns an `ActionEvent` — or `NULL` for "no
intervention."

### Policy A: always accept

The simplest possible policy — accept every dispatch regardless of state.
This is equivalent to having no policy at all (the baseline behavior).

```{r}
policy_always_accept <- list(

  propose_action = function(decision_point, entity, sim_ctx, param_ctx) {
    ActionEvent(
      action_type       = "accept",
      time_next         = entity$last_time + 0.01,
      decision_point_id = decision_point$id
    )
  }
)
```

### Policy B: battery threshold

Decline if battery has dropped below 25%. The idea: a low-battery agent
should stop taking new assignments and coast to shift end rather than
risk stranding mid-delivery.

```{r}
policy_battery_threshold <- list(
  propose_action = function(decision_point, entity, sim_ctx, param_ctx) {
    battery <- as.numeric(entity$current$battery_pct)

    action <- if (!is.null(battery) && battery < 25) "decline" else "accept"

    ActionEvent(
      action_type       = action,
      time_next         = entity$last_time + 0.01,
      decision_point_id = decision_point$id,
      metadata          = list(battery_at_decision = battery)
    )
  }
)
```

## Extending the bundle to handle the "decline" action

When the policy proposes "decline", that action enters the event timeline as
an `ActionEvent` and flows through `transition()`. We need the transition
function to handle it — in this case, by reverting the dispatch (resetting
mode back to idle and dropping the payload that was just assigned).

```{r}
# Extended transition that also handles the "decline" action
delivery_transition_with_policy <- function(entity, event, param_ctx = NULL) {
  # Handle the decline action: undo the dispatch assignment

  if (identical(event$event_type, "decline")) {
    return(list(
      dispatch_mode = "idle",
      payload_kg    = 0
    ))
  }

  # "accept" action: no additional state change needed (dispatch already applied)
  if (identical(event$event_type, "accept")) {
    return(NULL)
  }

  # All other events: delegate to the standard transition
  delivery_transition(entity, event, param_ctx)
}
```

Now build a bundle that uses this extended transition:

```{r}
delivery_bundle_with_policy <- function(params = list()) {
  base <- delivery_bundle(params)
  base$transition <- delivery_transition_with_policy
  base$event_catalog <- c(base$event_catalog, "accept", "decline")
  base
}
```

## Assembling with `load_model()`

When decision points are present, `load_model()` is the recommended assembly
path. It validates that the schema, bundle, policy, and trajectory configuration
are mutually consistent before any run begins.

```{r}
model_accept <- load_model(
  schema     = schema_with_dp,
  bundle     = delivery_bundle_with_policy(),
  policy     = policy_always_accept,
  trajectory = list(detail = "summary")
)

model_threshold <- load_model(
  schema     = schema_with_dp,
  bundle     = delivery_bundle_with_policy(),
  policy     = policy_battery_threshold,
  trajectory = list(detail = "summary")
)
```

## Running both policies from the same seed

The power of the decision-point architecture: same entity, same seed, same
stochastic draws — but different policies yield different outcomes.

```{r}
agent <- Entity$new(
  id          = "driver_A",
  init        = list(
    battery_pct   = 80,
    route_zone    = "urban",
    payload_kg    = 0,
    dispatch_mode = "idle"
  ),
  schema      = delivery_schema(),
  entity_type = "delivery_agent",
  time0       = 0
)

# Run under always-accept
out_accept <- model_accept$run(agent, max_events = 500,
                               return_observations = TRUE, seed = 99)

# Fresh entity (same starting state)
agent2 <- Entity$new(
  id          = "driver_A",
  init        = list(
    battery_pct   = 80,
    route_zone    = "urban",
    payload_kg    = 0,
    dispatch_mode = "idle"
  ),
  schema      = delivery_schema(),
  entity_type = "delivery_agent",
  time0       = 0
)

# Run under battery-threshold
out_threshold <- model_threshold$run(agent2, max_events = 500,
                                     return_observations = TRUE, seed = 99)
```

## Comparing outcomes

```{r}
# Count deliveries completed under each policy
count_deliveries <- function(out) {
  sum(out$events$event_type == "delivery_completed", na.rm = TRUE)
}

cat("Always-accept policy:\n")
cat("  Deliveries completed:", count_deliveries(out_accept), "\n")
cat("  Final battery:       ", round(out_accept$entity$current$battery_pct, 1), "%\n\n")

cat("Battery-threshold policy:\n")
cat("  Deliveries completed:", count_deliveries(out_threshold), "\n")
cat("  Final battery:       ", round(out_threshold$entity$current$battery_pct, 1), "%\n")
```

The always-accept agent takes every dispatch and drains the battery more
aggressively — potentially completing more deliveries but at higher risk of
hitting critical battery levels. The threshold agent conserves energy by
declining late-shift dispatches when battery is low.

## Inspecting trajectory records

Every time a decision point fires, the engine emits a `TrajectoryRecord`
capturing what was observed, what was proposed, and what happened next. This
is the audit trail that enables policy comparison and counterfactual analysis.

```{r}
# How many decisions were made?
cat("Decisions (accept policy):   ", length(out_accept$trajectory_records), "\n")
cat("Decisions (threshold policy):", length(out_threshold$trajectory_records), "\n")
```

Inspect the structure of a single record:

```{r}
tr <- out_threshold$trajectory_records[[1]]
cat("Time:            ", tr$t, "\n")
cat("Decision point:  ", tr$decision_point_id, "\n")
cat("Action proposed: ", tr$selected_action$action_type, "\n")
cat("Battery before:  ", tr$state_before$battery_pct, "\n")
cat("Battery after:   ", tr$state_after$battery_pct, "\n")
```

Build a summary table across all records:

```{r}
tr_df <- trajectory_table(out_threshold$trajectory_records,
                          vars = c("battery_pct", "dispatch_mode"))
head(tr_df, 10)
```

Look for the moment the policy starts declining — that's where the battery
crosses the threshold and the agent switches from "accept everything" to
"coast to end of shift."

## Cohort-level comparison

Let's scale this up. Run a 20-agent cohort under both policies and compare
aggregate outcomes.

```{r}
make_agents <- function(n = 20, seed = 2026) {
  set.seed(seed)
  lapply(seq_len(n), function(i) {
    Entity$new(
      id   = paste0("driver_", sprintf("%02d", i)),
      init = list(
        battery_pct   = runif(1, 50, 100),
        route_zone    = sample(c("urban", "suburban", "rural"), 1,
                               prob = c(0.55, 0.30, 0.15)),
        payload_kg    = 0,
        dispatch_mode = "idle"
      ),
      schema      = delivery_schema(),
      entity_type = "delivery_agent",
      time0       = 0
    )
  })
}

cohort_accept    <- lapply(make_agents(), function(e) {
  model_accept$run(e, max_events = 500, return_observations = TRUE, seed = 42)
})

cohort_threshold <- lapply(make_agents(), function(e) {
  model_threshold$run(e, max_events = 500, return_observations = TRUE, seed = 42)
})

# Aggregate
del_accept    <- vapply(cohort_accept, count_deliveries, integer(1))
del_threshold <- vapply(cohort_threshold, count_deliveries, integer(1))
bat_accept    <- vapply(cohort_accept,
                        function(o) o$entity$current$battery_pct, numeric(1))
bat_threshold <- vapply(cohort_threshold,
                        function(o) o$entity$current$battery_pct, numeric(1))

cat("Fleet summary — always accept:\n")
cat("  Mean deliveries:", round(mean(del_accept), 1), "\n")
cat("  Mean final battery:", round(mean(bat_accept), 1), "%\n\n")

cat("Fleet summary — battery threshold:\n")
cat("  Mean deliveries:", round(mean(del_threshold), 1), "\n")
cat("  Mean final battery:", round(mean(bat_threshold), 1), "%\n")
```

The trade-off is visible at fleet scale: the threshold policy sacrifices some
delivery throughput in exchange for better battery preservation — a real
operational consideration when battery replacement or charging infrastructure
is constrained.

## What trajectory records enable

`TrajectoryRecord` is not just an audit log. It is the structured surface that
the rest of the flux ecosystem builds on:

- **Policy comparison**: run the same entity under two policies with the same
  seed; diff the trajectory_records to see where and why they diverge.
- **Counterfactual analysis**: fix seed + parameter draw; vary only the policy;
  compare outcomes.
- **RL training** (future: `fluxSim`): the `(observation, action, reward,
  next_state)` tuple for each record becomes a training transition.
- **Audit and explainability**: for any individual run, reconstruct the full
  decision history and ask "why did the simulation do that?"

## Summary

| Concept | What you learned |
|---------|-----------------|
| `DecisionPoint()` | Declares where in the event timeline a policy is consulted |
| Policy function | `propose_action(dp, entity, sim_ctx, param_ctx)` → `ActionEvent` or NULL |
| `ActionEvent()` | The proposed intervention — enters the timeline like any other event |
| `load_model()` | Validates schema + bundle + policy + trajectory config together |
| Trajectory records | Per-decision audit trail: observation, action, state_before/after |
| Same seed, different policy | Isolates the causal effect of the policy on outcomes |

**Next:** [05_prepare_operational_data.md](05_prepare_operational_data.md) —
generate synthetic operational logs and prepare them into train/test/validation
format with `fluxPrepare`.
