---
title: "Validation"
maturity: "draft"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
suppressPackageStartupMessages({
  library(fluxCore)
  library(fluxPrepare)
  library(fluxForecast)
  library(fluxValidation)
})
```

A model that can simulate is not automatically a model you should trust. The
delivery model from Tutorial 03 produces convincing-looking trajectories, but
until you've tested its predictions against held-out data, you don't know
whether its rates, distributions, and state dynamics actually match reality.

This tutorial closes the loop: we take the test-set intervals built in Tutorial
05, forecast from each agent's reconstructed state, and compare predictions
against observed outcomes. By the end you will be able to:

- reconstruct test-set entities at their forecast anchor times,
- generate probabilistic forecasts from those baselines,
- build an observed outcome grid for comparison,
- run `validate_event_risk()` to assess calibration and discrimination,
- run `validate_state_point()` to compare predicted vs observed state.

## Setup and data preparation

We regenerate the same data pipeline from Tutorial 05. In practice you'd save
the TTV to disk; here we re-run for self-containedness.

```{r}
source("tutorials/model/urban_delivery.R")
source("tutorials/model/urban_delivery_data.R")

set.seed(42)
ops <- generate_delivery_log(n_agents = 30, n_shifts = 8)

splits <- generate_splits(ops$entities, train_frac = 0.6, test_frac = 0.2,
                          seed = 123)

events_prep <- prepare_events(
  events    = ops$events,
  id_col    = "entity_id",
  time_col  = "time",
  type_col  = "event_type",
  time_spec = time_spec(unit = "hours")
)

battery_spec <- list(
  id_col   = "entity_id",
  time_col = "time",
  vars     = "battery_pct"
)

obs_prep <- prepare_observations(
  tables    = list(battery = ops$observations),
  specs     = list(battery = battery_spec),
  time_spec = time_spec(unit = "hours")
)

splits_prep <- prepare_splits(splits, id_col = "entity_id", split_col = "split")

delivery_ep_spec <- spec_event_process(
  event_types = "delivery_completed",
  name        = "delivery_completion",
  t0_strategy = "followup_start"
)

ttv <- build_ttv_event_process(
  events       = events_prep,
  observations = obs_prep,
  splits       = splits_prep,
  spec         = delivery_ep_spec,
  followup     = ops$followup,
  time_spec    = time_spec(unit = "hours")
)
```

## The test set

Each row in `ttv$test` represents one agent × one shift interval:

```{r}
head(ttv[ttv$split == "test", ])
cat("Test-set intervals:", sum(ttv$split == "test"), "\n")
cat("Outcome rate:      ", round(mean(ttv$event_occurred[ttv$split == "test"]), 3), "\n")
```

The `event_occurred` column is TRUE if the agent completed at least one delivery
during the interval. The outcome rate tells us how "easy" the prediction
task is — if it's close to 1.0, nearly everyone completes a delivery (which is
expected for an 8-hour shift).

## Reconstruct agent state at t₀

For each test-set interval, we need the agent's state at the forecast anchor
time. `reconstruct_state_at()` pulls the last observed battery_pct before t₀:

```{r}
ttv_test <- ttv[ttv$split == "test", ]

state_at_t0 <- reconstruct_state_at(
  anchors      = ttv_test[, c("entity_id", "t0")],
  observations = obs_prep,
  vars         = "battery_pct",
  id_col       = "entity_id",
  time_col     = "t0",
  time_spec    = time_spec(unit = "hours")
)

head(state_at_t0)
```

Now build Entity objects from these reconstructed states:

```{r}
shared_schema <- delivery_schema()

test_entities <- lapply(seq_len(nrow(state_at_t0)), function(i) {
  row <- state_at_t0[i, ]
  battery <- if (is.na(row$battery_pct)) 80 else row$battery_pct

  Entity$new(
    id          = row$entity_id,
    init = list(
      battery_pct   = battery,
      route_zone    = "urban",    # default; could be reconstructed too
      payload_kg    = 0,
      dispatch_mode = "idle"
    ),
    schema      = shared_schema,
    entity_type = "delivery_agent",
    time0       = row$t0
  )
})
names(test_entities) <- state_at_t0$entity_id
```

## Forecast from test-set baselines

Run the model forward from each reconstructed entity state:

```{r}
eng <- Engine$new(bundle = delivery_bundle())

# Evaluation times: hours 1 through 8 from each entity's t0
# For simplicity, use a fixed horizon grid relative to t0=0
horizon <- 8
times <- seq(1, horizon, by = 1)

fc <- forecast(
 engine   = eng,
  entities = test_entities,
  times    = times,
  S        = 100,
  seed     = 42
)
```

## Predicted event probabilities

"What does the model predict as the probability of delivery completion by
each horizon time?"

```{r}
ep <- event_prob(fc, event = "delivery_completed", times = times)
head(ep)
```

## Build the observed outcome grid

`build_obs_grid()` aligns the actual observed events onto the same time grid
the forecast was evaluated on. This creates the "ground truth" that predictions
are compared against.
```{r}
# Subset observations to test-set entities
test_ids <- unique(state_at_t0$entity_id)
test_obs_df <- as.data.frame(obs_prep[obs_prep$entity_id %in% test_ids,
                                       c("entity_id", "time", "battery_pct")])
test_events <- events_prep[events_prep$entity_id %in% test_ids, ]

obs_grid <- build_obs_grid(
  vars           = list(battery = test_obs_df),
  events         = test_events,
  times          = times,
  t0             = 0,
  start_time     = 0,
  time_spec      = time_spec(unit = "hours"),
  id_col         = "entity_id",
  time_col       = "time",
  event_time_col = "time",
  event_type_col = "event_type"
)
```

Each row in the obs_grid says: "for entity X, did event Y occur by time Z?"
This is the binary ground truth that `validate_event_risk()` compares against
the predicted probabilities.

## `validate_event_risk()` — calibration and discrimination

The core validation function compares predicted event probabilities against
observed binary outcomes:

```{r}
vr <- validate_event_risk(
  pred     = ep,
  obs      = obs_grid,
  event    = "delivery_completed",
  times    = times
)
```

The result contains calibration and discrimination metrics:

```{r}
str(vr, max.level = 1)
```

### Calibration

Calibration asks: "when the model predicts 70% probability, does the event
actually occur ~70% of the time?"

```{r}
if (!is.null(vr$calibration)) {
  print(vr$calibration)
}
```

Perfect calibration means the predicted probability matches the observed
frequency in each bin. Systematic over-prediction (predicted > observed) suggests
the model's rates are too aggressive; under-prediction suggests they're too
conservative.

### Discrimination

Discrimination asks: "can the model distinguish agents who will complete a
delivery from those who won't?"

```{r}
if (!is.null(vr$discrimination)) {
  cat("C-statistic:", round(vr$discrimination$c_stat, 3), "\n")
}
```

A C-statistic of 0.5 means no discrimination (random guessing); 1.0 means
perfect separation. For our synthetic data (where the model generated the data),
we expect strong discrimination. In real applications, this is where model
deficiencies show up.

## `validate_state_point()` — predicted vs observed state

Beyond event probabilities, we can compare the model's predicted battery
distribution against what was actually observed. This validates the state
dynamics, not just the event timing.

```{r}
# Get predicted state distribution
ss <- state_summary(fc, vars = "battery_pct", times = times)

# Get observed battery values from the test-set observations
test_ids <- unique(state_at_t0$entity_id)
test_obs <- obs_prep[obs_prep$entity_id %in% test_ids, ]

vs <- validate_state_point(
  pred       = ss,
  obs        = test_obs,
  var        = "battery_pct",
  times      = times,
  start_time = 0
)

if (!is.null(vs)) {
  str(vs, max.level = 1)
}
```

## Interpreting miscalibration

When validation reveals miscalibration, the model is telling you something
specific. For the delivery model, common failure modes:

| Observation | Likely cause | Fix |
|-------------|-------------|-----|
| Over-prediction of delivery events | `delivery_rate_base` too high | Lower the rate; check if real-world completion times are longer |
| Under-prediction early in shift | `dispatch_rate_base` too low for the first hour | Consider time-varying dispatch rates |
| Battery drains too fast in model | `delivery_battery_drop_mean` too high | Calibrate against real sensor data |
| Battery drains too slow | Missing energy-consuming processes (e.g., GPS, AC) | Add an ambient drain process |

The feedback loop: **observed outcomes → identify miscalibrated parameters →
update `delivery_bundle(params = ...)` → re-validate**. This is the core
model development cycle that flux supports.

## Summary

| Function | Purpose |
|----------|---------|
| `reconstruct_state_at()` | Recover agent state at forecast anchor from history |
| `forecast()` | Run model forward from test-set baselines |
| `event_prob()` | Predicted probability of named event by time *t* |
| `build_obs_grid()` | Align observed events onto evaluation time grid |
| `validate_event_risk()` | Compare predicted probabilities vs observed outcomes |
| `validate_state_point()` | Compare predicted state distribution vs observed |

This is the complete prediction–validation loop. The same pattern applies
regardless of domain: reconstruct state → forecast → compare → iterate.

**The tutorial suite so far:**

| # | Tutorial | Packages |
|---|----------|----------|
| 01 | Engine and ModelBundle scaffold | fluxCore |
| 03 | Cohort simulation and forecast | fluxCore, fluxForecast |
| 04 | Decision points and policy | fluxCore |
| 05 | Preparing operational data | fluxCore, fluxPrepare |
| 06 | Validation (this tutorial) | fluxCore, fluxPrepare, fluxForecast, fluxValidation |

Together these cover the full lifecycle: define a model → simulate cohorts →
add decisions → prepare real data → validate predictions. The ecosystem is
designed so each piece slots into this pipeline without custom glue code.
